---
title: "Untitled"
format: html
server: shiny
---

---
title: "Untitled"
format: html
server: shiny
---

In this Quarto document, we create four embedded MongoDB databases from four existing SQL DynLib databases. We transform the relational structure of the SQL databases into a collection-based NoSQL format, and finally, we benchmark queries on both types of databases.

```{r}
library(RSQLite)
library(mongolite)
library(jsonlite)
library(dplyr)
library(stringi)
library(DBI)
library(janitor)
library(microbenchmark)
pth <- getwd()
pth

```

```{r}
#define SQLite documents
sqlite_files <- c("data/Dylib_FTMS_neg_or.sqlite", 
                  "data/Dylib_FTMS_pos2.sqlite",
                  "data/Dylib_QTOF_neg.sqlite", 
                  "data/Dylib_QTOF_pos.sqlite")

#giving names for the resulting MongoDb databases
mongo_dbs <- c("Dylib_FTMS_neg_mongo2", "Dylib_FTMS_pos_mongo2",
               "Dylib_QTOF_neg_mongo2", "Dylib_QTOF_pos_mongo2")

#giving names for the resulting MongoDb collections
mongo_collections <- c("coll_FTMS_neg2", "coll_FTMS_pos2", 
                       "coll_QTOF_neg2", "coll_QTOF_pos2")

```

```{r}

# Cleaning columns
clean_utf8 <- function(df) {
  df[] <- lapply(df, function(col) {
    if (is.character(col)) stri_encode(col, "", "UTF-8") else col
  })
  return(df)
}

```

```{r}
# Create a folder for saving the resulting json files
json_dir <- "data/embaded2_json"
if (!dir.exists(json_dir)) {
  dir.create(json_dir, recursive = TRUE)
}
```

```{r}
# Loop over SQLite files
for (i in seq_along(sqlite_files)) {
  sqlite_file <- sqlite_files[i]
  mongo_db <- mongo_dbs[i]
  mongo_collection <- mongo_collections[i]

  message("\nProcessing ", basename(sqlite_file))

  con <- dbConnect(SQLite(), sqlite_file)

  # Read & clean tables
  experiment_tbl <- dbReadTable(con, "experiment") %>% clean_utf8()
  msms_spectrum_tbl <- dbReadTable(con, "msms_spectrum") %>% clean_utf8()
  msms_peak_tbl <- dbReadTable(con, "msms_spectrum_peak") %>% clean_utf8()
  compound_tbl <- dbReadTable(con, "ms_compound") %>% clean_utf8()

  dbDisconnect(con)

  # Insert experiment collection
  mongo_exp <- mongo(collection = "experiment", db = mongo_db)
  mongo_exp$remove('{}')
  mongo_exp$insert(experiment_tbl)
  message("Experiment collection inserted with ", nrow(experiment_tbl), " documents")

  # Group peaks and compounds
  peaks_by_spec <- split(msms_peak_tbl, msms_peak_tbl$spectrum_id)
  compound_by_id <- split(compound_tbl, compound_tbl$compound_id)

  # Build spectrum documents
  link_docs <- lapply(seq_len(nrow(msms_spectrum_tbl)), function(idx) {
    spec_row <- as.list(msms_spectrum_tbl[idx, , drop = FALSE])
    spec_id <- spec_row$spectrum_id
    cmp_id <- spec_row$compound_id

    # Structure peaks as array of objects
    peaks <- peaks_by_spec[[as.character(spec_id)]]
    if (!is.null(peaks)) {
      spec_row$peaks <- lapply(seq_len(nrow(peaks)), function(j) {
        list(mz = as.numeric(peaks$mz[j]), intensity = as.numeric(peaks$intensity[j]))
      })
    } else {
      spec_row$peaks <- list()
    }

    # Array of compound objects
    compound <- compound_by_id[[as.character(cmp_id)]]
    if (!is.null(compound)) {
      spec_row$compound <- lapply(seq_len(nrow(compound)), function(j) {
        as.list(compound[j, , drop = FALSE])
      })
    } else {
      spec_row$compound <- list()
    }

    return(spec_row)
  })

  #'Insert all documents into MongoDB 
  #'and write a JSON file for each database
  mongo_conn <- mongo(collection = mongo_collection, db = mongo_db)
  mongo_conn$remove('{}')

  json_docs <- vapply(
    link_docs,
    FUN = function(doc) {
      tryCatch(
        toJSON(doc, auto_unbox = TRUE, null = "null", POSIXt = "ISO8601", na = "null"),
        error = function(e) {
          warning("Invalid JSON: ", conditionMessage(e))
          NA_character_
        }
      )
    },
    FUN.VALUE = character(1)
  )

  valid_docs <- json_docs[!is.na(json_docs)]

  # Save one JSON file per SQLite database
  json_file <- file.path(json_dir, paste0(tools::file_path_sans_ext(basename(sqlite_file)), ".json"))
  writeLines(valid_docs, con = json_file)
  message("Saved JSON file: ", json_file)

  # Insert into MongoDB
  if (length(valid_docs) > 0) {
    mongo_conn$insert(valid_docs, parse = TRUE)
    message("Inserted ", length(valid_docs), " documents into MongoDB.")
  } else {
    message("No valid documents to insert into MongoDB.")
  }

  # Add indexes for performance
  mongo_conn$index(add = '{"spectrum_id": 1}')
  mongo_conn$index(add = '{"peaks.intensity": 1}')

  message("MongoDB import & indexing done for DB: ", mongo_db, " / Collection: ", mongo_collection)
}

```

**Data Manipulation**

```{r}
library(mongolite)
#Connexion to the Dylib_FTMS_neg_or database
coll <- mongo(collection = "coll_FTMS_neg2", db = "Dylib_FTMS_neg_mongo2")
data <- coll$find()
head(data)

```

```{r}
# Count the number of document in ms_compound collection (201034)
nb_documents <- coll$count()
nb_documents 
```

```{r}
# Print the first document in compound field
doc <- coll$find(limit = 1)
print(doc$compound)
```

```{r}
# Print the first document in peaks field
doc <- coll$find(limit = 1)
print(doc$peaks)

```

```{r}
# find the documents where compound.expid field is equal to 18
result <- coll$find('{"compound.expid": 18}', limit = 4)
print(result)
```

```{r}
library(mongolite)
#Connexion to the experiment collection of the Dylib_FTMS_neg_or database
coll_exp <- mongo(collection = "experiment", db = "Dylib_FTMS_neg_mongo2")
data <- coll_exp$find()
head(data)


```

```{r}
# Count the number of document in ms_compound collection (265)
nb_doc_exp <- coll_exp$count()
nb_doc_exp 

```

```{r}
# Get the first document
# Print the first document in experiment collection
doc_exp <- coll_exp$find(limit = 1)
print(doc_exp)

```

```{r}
# find the documents where expid field is equal to 18
exp18 <- coll_exp$find('{"expid": 18}', limit = 4)
print(exp18)
```

```{r}
#Connexion to the experiment collection of the Dylib_FTMS_neg_or database
coll_qpos <- mongo(collection = "coll_QTOF_pos2", db = "Dylib_QTOF_pos_mongo2")
data <- coll_qpos$find()
head(data)
```

```{r}
# Print the first document in compound field
doc_qpos <- coll_qpos$find(limit = 1)
print(doc_qpos$compound)
```

```{r}
# Count the number of document in ms_compound collection (6237)
nb_doc_qpos <- coll_qpos$count()
nb_doc_qpos 
```

```{r}
# Print the first document in peaks field
doc_pk_qpos <- coll_qpos$find(limit = 1)
print(doc_pk_qpos$peaks)
```

```{r}
# Print the first document in compound field
doc_cmp_qpos <- coll_qpos$find(limit = 1)
print(doc_cmp_qpos$compound)
```

**Performance Test between SQL and MongoDB:**

First, we compare the time performance of a simple query that extracts documents with specific `spectrum_id`s from both SQL and MongoDB databases.

```{r}

# Spectrum IDs - numeric
spectrum_ids <- c(1, 5, 8)

# SQLite connection
sqlite_con <- dbConnect(RSQLite::SQLite(), "data/DyLib_QTOF_neg.sqlite")

# MongoDB connection
mongo_con <- mongo(collection = "coll_QTOF_neg", db = "Dylib_QTOF_neg_mongo")

# SQLite extraction function
get_from_sqlite <- function(ids) {
  dbGetQuery(sqlite_con, sprintf(
    "SELECT * FROM msms_spectrum WHERE spectrum_id IN (%s)",
    paste(ids, collapse = ", ")
  ))
}


gget_from_mongo <- function(ids) {
  pipeline <- sprintf('[
                      {
                      "$match" : {
                         "spectrum_id" : {"$in" : [%s]}
                      }
                      }
                      
                      ]',paste(ids, collapse = ","))
   mongo_con$aggregate(pipeline)
}



print(get_from_mongo(spectrum_ids))  # This should return documents

# Benchmarking
benchmark_result <- microbenchmark(
  sqlite = invisible(get_from_sqlite(spectrum_ids)),
  mongo  = invisible(get_from_mongo(spectrum_ids)),
  times = 20L
)

# Results
print(benchmark_result)
boxplot(benchmark_result)


```

MongoDB query takes much more time for execution then SQL that was super fast for extracting documents for those specific 3 spectrum_id.

Next we use another query to extract peaks with an intensity greater than 100 for a given spectrum_ids

```{r}
# Spectrum IDs to query
spectrum_ids <- c(1, 5, 8)

# SQLite connection
sqlite_con <- dbConnect(RSQLite::SQLite(), "data/DyLib_QTOF_neg.sqlite")

# MongoDB connection
mongo_con <- mongo(collection = "coll_QTOF_neg", db = "Dylib_QTOF_neg_mongo")

# SQLite retrieval function
get_from_sqlite <- function(ids) {
  query <- sprintf(
    "SELECT spectrum_id, mz, intensity FROM msms_spectrum_peak WHERE spectrum_id IN (%s) AND intensity > 100",
    paste(ids, collapse = ", ")
  )
  dbGetQuery(sqlite_con, query)
}

get_from_mongo_agg <- function(ids) {
  ids_str <- paste(ids, collapse = ", ")
  
  pipeline <- sprintf('[
              
    {
      "$match": {
        "spectrum_id": { "$in": [%s] },
        "peaks": { "$elemMatch": { "intensity": { "$gt": 100 } } }
      }
    },
    {
      "$project": {
        "spectrum_id": 1,
        "peaks": 1,
        "_id": 0
      }
    }
  ]', ids_str)
  
  mongo_con$aggregate(pipeline)
}




# Run benchmark
benchmark_result <- microbenchmark(
  sqlite = invisible(get_from_sqlite(spectrum_ids)),
  mongo  = invisible(get_from_mongo(spectrum_ids)),
  times = 20L
)

# Show results
print(benchmark_result)
boxplot(benchmark_result, main = "SQLite vs MongoDB Query Time", ylab = "Time (microseconds)")

```

Also for peaks extraction SQL was fast compared to MongoDB query.

Next we use another query to extract peaks with an intensity greater than 100 for 1000 spectrum_ids

```{r}
library(DBI)
library(RSQLite)
library(mongolite)
library(microbenchmark)
library(dplyr)

# CONNECTIONS 
sqlite_con <- dbConnect(SQLite(), "data/DyLib_QTOF_neg.sqlite")
mongo_con <- mongo(collection = "coll_QTOF_neg", db = "Dylib_QTOF_neg_mongo")

# RETRIEVE 1000 SPECTRUM IDs FROM SQLITE 
spectrum_ids <- dbGetQuery(sqlite_con, "SELECT DISTINCT spectrum_id FROM msms_spectrum_peak LIMIT 10000")$spectrum_id

# SQLITE QUERY FUNCTION
get_from_sqlite <- function(ids) {
  id_str <- paste(ids, collapse = ", ")
  query <- sprintf(
    "SELECT spectrum_id, mz, intensity FROM msms_spectrum_peak WHERE spectrum_id IN (%s) AND intensity > 100",
    id_str
  )
  dbGetQuery(sqlite_con, query)
}

# MONGODB QUERY FUNCTION
get_from_mongo_filtered_peaks <- function(ids) {
  # Use jsonlite to format the array safely
  id_json <- toJSON(ids, auto_unbox = TRUE)
  
  pipeline <- sprintf('[ 
    { "$match": { "spectrum_id": { "$in": %s } } },
    { "$unwind": "$peaks" },
    { "$match": { "peaks.intensity": { "$gt": 100 } } },
    { "$project": {
        "spectrum_id": 1,
        "mz": "$peaks.mz",
        "intensity": "$peaks.intensity",
        "_id": 0
    }}
  ]', id_json)

  mongo_con$aggregate(pipeline)
}

# RUN BENCHMARK 
benchmark_result <- microbenchmark(
  sqlite = invisible(get_from_sqlite(spectrum_ids)),
  mongo  = invisible(get_from_mongo_filtered_peaks(spectrum_ids)),
  times = 10L
)

print(benchmark_result)
boxplot(benchmark_result, main = "SQLite vs MongoDB (Filtered Peaks)", ylab = "Time (ms)")


```

MongoDB here is faster than SQL

In this script we select rows/documents with an mz between 100 and 200 and intensity \> 5000

```{r}
#CONNECTIONS
sqlite_con <- dbConnect(SQLite(), "data/DyLib_QTOF_neg.sqlite")
mongo_con <- mongo(collection = "coll_QTOF_neg", db = "Dylib_QTOF_neg_mongo")

# CREATE COMPOUND MULTIKEY INDEX ON MONGODB 
# This index helps MongoDB quickly find peaks matching both mz and intensity criteria
mongo_con$index(add = '{"peaks.mz": 1, "peaks.intensity": 1}')

# SQLITE QUERY FUNCTION 
get_peaks_from_sqlite_filtered <- function() {
  query <- "
    SELECT spectrum_id, mz, intensity 
    FROM msms_spectrum_peak 
    WHERE mz BETWEEN 100 AND 200 AND intensity > 5000
  "
  dbGetQuery(sqlite_con, query)
}

# MONGODB AGGREGATION FUNCTION WITH OPTIMIZED FILTERING 
get_peaks_from_mongo_filtered <- function() {
  pipeline <- '[
    { "$match": { 
        "peaks": { 
          "$elemMatch": { 
            "mz": { "$gte": 100, "$lte": 200 }, 
            "intensity": { "$gt": 5000 }
          }
        }
      }
    },
    { "$unwind": "$peaks" },
    { "$match": {
        "peaks.mz": { "$gte": 100, "$lte": 200 },
        "peaks.intensity": { "$gt": 5000 }
      }
    },
    { "$project": {
        "spectrum_id": 1,
        "mz": "$peaks.mz",
        "intensity": "$peaks.intensity",
        "_id": 0
      }
    }
  ]'
  mongo_con$aggregate(pipeline)
}

# RUN BENCHMARK 
benchmark_result <- microbenchmark(
  sqlite = invisible(get_peaks_from_sqlite_filtered()),
  mongo  = invisible(get_peaks_from_mongo_filtered()),
  times = 10L
)

# RESULTS 
print(benchmark_result)
boxplot(benchmark_result, main = "SQLite vs MongoDB (Filtered Peaks mz 100-200, intensity > 5000)", ylab = "Time (ms)")

```

The benchmark results show that mongodb is faster than SQL.

Here we just changed the filtering criteria to return fewer peaks by selecting rows/documents that have mz between 200 and 210 and intensity \> 10000

```{r}
library(DBI)
library(RSQLite)
library(mongolite)
library(microbenchmark)

# Connections
sqlite_con <- dbConnect(SQLite(), "data/DyLib_QTOF_neg.sqlite")
mongo_con <- mongo(collection = "coll_QTOF_neg", db = "Dylib_QTOF_neg_mongo")

# Create compound multikey index in MongoDB for peaks.mz and peaks.intensity
mongo_con$index(add = '{"peaks.mz": 1, "peaks.intensity": 1}')

# SQLite function: query peaks with mz between 200 and 210 and intensity > 10000
get_peaks_sqlite_filtered <- function() {
  query <- "
    SELECT spectrum_id, mz, intensity
    FROM msms_spectrum_peak
    WHERE mz BETWEEN 200 AND 210 AND intensity > 10000
  "
  dbGetQuery(sqlite_con, query)
}

# MongoDB function: optimized aggregation with early $match on nested fields
get_peaks_mongo_filtered <- function() {
  pipeline <- '[
    { "$match": {  
        "peaks": {
          "$elemMatch": {
            "mz": { "$gte": 200, "$lte": 210 },
            "intensity": { "$gt": 10000 }
          }
        }
      }
    },
    { "$unwind": "$peaks" },
    { "$match": {
        "peaks.mz": { "$gte": 200, "$lte": 210 },
        "peaks.intensity": { "$gt": 10000 }
      }
    },
    { "$project": {
        "spectrum_id": 1,
        "mz": "$peaks.mz",
        "intensity": "$peaks.intensity",
        "_id": 0
      }
    }
  ]'
  mongo_con$aggregate(pipeline)
}

# Benchmark
benchmark_result <- microbenchmark(
  sqlite = invisible(get_peaks_sqlite_filtered()),
  mongo  = invisible(get_peaks_mongo_filtered()),
  times = 10L
)

print(benchmark_result)
boxplot(benchmark_result, main = "SQLite vs MongoDB: Peaks mz 200-210 & intensity > 10000", ylab = "Time (ms)")


```

Also MongoDB is faster than SQLite in this benchmark, As a result mongoDB can outperform SQL for highly selective queries on deeply nested data.

Here we group by spectrum_id and we select peaks where mz is greater or equal than 3000 and intensity is greater than 10000

```{r}

mongo_con$index(add = '{"peaks.mz": 1, "peaks.intensity": 1}')

get_nested_summary_mongo <- function() {
  pipeline <- '[
  { "$match": {  
        "peaks": {
          "$elemMatch": {
            "mz": { "$gte": 3000},
            "intensity": { "$gt": 10000 }
          }
        }
      }
    },
    { "$unwind": "$peaks" },
    { "$match": {
        "peaks.mz": { "$gt": 3000 },
        "peaks.intensity": { "$gt": 10000 }
    }},
    { "$group": {
        "_id": "$spectrum_id",
        "n_peaks": { "$sum": 1 }
    }}
  ]'
  mongo_con$aggregate(pipeline)
}


 

get_nested_summary_sqlite <- function() {
  query <- "
    SELECT spectrum_id,
           COUNT(*) AS num_peaks,
           AVG(intensity) AS avg_intensity
    FROM msms_spectrum_peak
    WHERE mz > 300 AND intensity > 1000
    GROUP BY spectrum_id
  "
  dbGetQuery(sqlite_con, query)
}

benchmark_result <- microbenchmark(
  mongo  = invisible(get_nested_summary_mongo()),
  sqlite = invisible(get_nested_summary_sqlite()),
  times = 10L
)

print(benchmark_result)
boxplot(benchmark_result, main = "MongoDB vs SQLite: Nested Aggregation", ylab = "Time (ms)")


```

the benchmark results show that mongodb is faster than SQLite for this query which ensure that mongoDB can outperform SQL for highly selective queries on deeply nested data.
