---
title: "Add the grouping and annotation columns to the sql database"
format: html
server: shiny
author: "Ahlam Mentag"
---

## Shiny Documents

This Quarto document is made interactive using Shiny. Interactive documents allow readers to modify parameters and see the results immediately. Learn more about Shiny interactive documents at <https://quarto.org/docs/interactive/shiny/>.

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change. This demonstrates how a standard R plot can be made interactive:

```{r}
library(DBI)
library(RSQLite)
pth <- getwd()
setwd(pth)
```

```{r}
FT_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/Similarity/FTMS/data/DYL_FLAX_FTMSneg.sqlite"
QTOF_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/Similarity/QTOF/data/DYL_FLAX_QTOFneg.sqlite"
assoc_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/New_Dynlib/data/combined_assoc.txt"
FT_con <- DBI::dbConnect(RSQLite::SQLite(), FT_path)
QTOF_con <- DBI::dbConnect(RSQLite::SQLite(), QTOF_path)
```


```{r}
dbListTables(FT_con)
```

```{r}
dbListFields(FT_con, "ms_compound")
```

```{r}
dbGetQuery(FT_con, "Select * from ms_compound Limit 3 ")
```

Scan all experiments to detect required columns
```{r}
base_dir <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/inst/database_FTMS_neg/Experimenten"

xcms_files <- list.files(
  base_dir,
  pattern = "xcms2.txt$",
  recursive = TRUE,
  full.names = TRUE
)

grp_cols_all <- character()
anno_cols_all <- character()

for (f in xcms_files) {
  x <- read.delim(f, nrows = 1, check.names = FALSE)

  grp_cols_all  <- union(grp_cols_all,
                          grep("^nm\\.Grpo\\.Grp", names(x), value = TRUE))
  anno_cols_all <- union(anno_cols_all,
                          grep("^anno", names(x), value = TRUE))
}

grp_cols_all
anno_cols_all


```

```{r}
existing_cols <- dbListFields(FT_con, "ms_compound")

add_if_missing <- function(colname) {
  if (!colname %in% existing_cols) {
    message("Adding column: ", colname)
    dbExecute(
      FT_con,
      sprintf("ALTER TABLE ms_compound ADD COLUMN `%s` TEXT", colname)
    )
  }
}

invisible(lapply(grp_cols_all, add_if_missing))
invisible(lapply(anno_cols_all, add_if_missing))

```


```{r}
base_dir <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/inst/database_FTMS_neg/Experimenten"
folders <- list.dirs(base_dir, recursive = FALSE)

```

```{r}
grp_cols_all  <- character()
anno_cols_all <- character()

for (f in xcms_files) {
  x <- read.delim(f, nrows = 1, check.names = FALSE)

  grp_cols_all  <- union(
    grp_cols_all,
    grep("^nm\\.Grpo\\.Grp", names(x), value = TRUE)
  )

  anno_cols_all <- union(
    anno_cols_all,
    grep("^anno", names(x), value = TRUE)
  )
}

grp_cols_all
anno_cols_all


```

```{r}
dbExecute(FT_con, "
CREATE INDEX IF NOT EXISTS idx_ms_compound_lookup
ON ms_compound (expid, nodename, retention_time, mass_measured)
")

```



```{r}
extract_expid <- function(folder_name) {
  as.integer(sub(".*Exp([0-9]+).*", "\\1", folder_name))
}

```

```{r}
build_ms_compound_df <- function(xcms, expid, grp_cols_all, anno_cols_all) {

  df <- data.frame(
    nodename = xcms$name,
    mass_measured = xcms$mzmed,
    retention_time = xcms$rtmed,
    expid = expid,
    stringsAsFactors = FALSE
  )

  for (cl in grp_cols_all) {
    df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
  }

  for (cl in anno_cols_all) {
    df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
  }

  df
}

```



```{r}
folders <- list.dirs(base_dir, recursive = FALSE)

for (fld in folders) {

  expid <- extract_expid(basename(fld))
  if (is.na(expid)) next

  xcms_file <- file.path(fld, "xcms2.txt")
  if (!file.exists(xcms_file)) next

  message("Processing experiment ", expid)

  xcms <- read.delim(
    xcms_file,
    check.names = FALSE,
    stringsAsFactors = FALSE
  )

  new_df <- build_ms_compound_df(
    xcms, expid, grp_cols_all, anno_cols_all
  )

  dbWithTransaction(FT_con, {

    ## ---- remove rows that already exist ----
    existing <- dbGetQuery(
      FT_con,
      "
      SELECT nodename, retention_time, mass_measured
      FROM ms_compound
      WHERE expid = ?
      ",
      params = list(expid)
    )

    if (nrow(existing) > 0) {

      key_existing <- paste(
        existing$nodename,
        round(existing$retention_time, 3),
        round(existing$mass_measured, 6)
      )

      key_new <- paste(
        new_df$nodename,
        round(new_df$retention_time, 3),
        round(new_df$mass_measured, 6)
      )

      new_df <- new_df[!key_new %in% key_existing, ]
    }

    ## ---- BULK INSERT ----
    if (nrow(new_df) > 0) {
      dbAppendTable(FT_con, "ms_compound", new_df)
    }
  })
}


```


```{r}
dbGetQuery(
  FT_con,
  "
  SELECT nodename, expid, `nm.Grpo.Grp`, `nm.Grpo.Grp.1`, anno
  FROM ms_compound
  WHERE expid = 254
  LIMIT 10
  "
)

```

