---
title: "Add the grouping and annotation columns to the sql database"
format: html
server: shiny
author: "Ahlam Mentag"
---

## Shiny Documents

In this quarto document we add the grouping and annotation columns to the new 
sql format of Dynlib FTMS and QTOF positive databases. 


```{r}
library(DBI)
library(RSQLite)
pth <- getwd()
setwd(pth)
```

Create sql connections to the FTMS and QTOF positive databases
```{r}
FT_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/data/DyLib_FTMS_pos.sqlite"
QTOF_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/data/DyLib_QTOF_pos.sqlite"
FT_con <- DBI::dbConnect(RSQLite::SQLite(), FT_path)
QTOF_con <- DBI::dbConnect(RSQLite::SQLite(), QTOF_path)
```

List tables of FTMS positive data
```{r}
dbListTables(FT_con)
```
List columns names of the ms_compound table
```{r}
dbListFields(FT_con, "ms_compound")
```

```{r}
dbGetQuery(FT_con, "Select * from ms_compound Limit 3 ")
```

Scan all experiments to detect required columns
```{r}
base_dir <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/inst/database_FTMS_pos/Experimenten"

xcms_files <- list.files(
  base_dir,
  pattern = "xcms2.txt$",
  recursive = TRUE,
  full.names = TRUE
)

grp_cols_all <- character()
anno_cols_all <- character()

for (f in xcms_files) {
  x <- read.delim(f, nrows = 1, check.names = FALSE)

  grp_cols_all  <- union(grp_cols_all,
                          grep("^nm\\.Grpo\\.Grp", names(x), value = TRUE))
  anno_cols_all <- union(anno_cols_all,
                          grep("^anno", names(x), value = TRUE))
}

grp_cols_all
anno_cols_all


```

```{r}
# Combine all columns you might add
all_cols <- c(grp_cols_all, anno_cols_all, extra_cols)

existing_cols <- dbListFields(FT_con, "ms_compound")

add_if_missing <- function(colname) {
  if (!colname %in% existing_cols) {
    message("Adding column: ", colname)
    dbExecute(
      FT_con,
      sprintf("ALTER TABLE ms_compound ADD COLUMN `%s` TEXT", colname)
    )
  }
}

# Add all missing columns
invisible(lapply(all_cols, add_if_missing))

```


```{r}
base_dir <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/inst/database_FTMS_pos/Experimenten"
folders <- list.dirs(base_dir, recursive = FALSE)

```

```{r}
grp_cols_all  <- character()
anno_cols_all <- character()

for (f in xcms_files) {
  x <- read.delim(f, nrows = 1, check.names = FALSE)

  grp_cols_all  <- union(
    grp_cols_all,
    grep("^nm\\.Grpo\\.Grp", names(x), value = TRUE)
  )

  anno_cols_all <- union(
    anno_cols_all,
    grep("^anno", names(x), value = TRUE)
  )
}

grp_cols_all
anno_cols_all


```

```{r}
dbExecute(FT_con, "
CREATE INDEX IF NOT EXISTS idx_ms_compound_lookup
ON ms_compound (expid, nodename, retention_time, mass_measured)
")

```


Extracting the experiment number from each experiment folder in the old dynlib
database
```{r}
extract_expid <- function(folder_name) {
  as.integer(sub(".*Exp([0-9]+).*", "\\1", folder_name))
}

```

```{r}
build_ms_compound_df <- function(xcms, expid, grp_cols_all = NULL, anno_cols_all = NULL, extra_cols = NULL) {

  df <- data.frame(
    nodename       = xcms$name,
    mass_measured  = xcms$mzmed,
    retention_time = xcms$rtmed,
    expid          = expid,
    stringsAsFactors = FALSE
  )

  # add grouping columns dynamically if present
  if (!is.null(grp_cols_all)) {
    for (cl in grp_cols_all) {
      df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
    }
  }

  # add annotation columns dynamically if present
  if (!is.null(anno_cols_all)) {
    for (cl in anno_cols_all) {
      df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
    }
  }

  # add extra metadata columns dynamically if present
  if (!is.null(extra_cols)) {
    for (cl in extra_cols) {
      df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
    }
  }

  df
}


```


Extracting the grouping and annotation columns from the "xcms2.txt" file of each
experiment:

```{r}
folders <- list.dirs(base_dir, recursive = FALSE)

for (fld in folders) {

  expid <- extract_expid(basename(fld))
  if (is.na(expid)) next

  xcms_file <- file.path(fld, "xcms2.txt")
  if (!file.exists(xcms_file)) next

  message("Processing experiment ", expid)

  xcms <- read.delim(
    xcms_file,
    check.names = FALSE,
    stringsAsFactors = FALSE
  )

# detect extra columns you want to include
extra_cols <- c("isf", "MeAb", "CorrPkGrp", "massMS1", "Comb",
                "con", "CON.new", "subgrp", "CON.new.1")

new_df <- build_ms_compound_df(
  xcms, expid, grp_cols_all, anno_cols_all, extra_cols
)


  dbWithTransaction(FT_con, {

    ##remove rows that already exist
    existing <- dbGetQuery(
      FT_con,
      "
      SELECT nodename, retention_time, mass_measured
      FROM ms_compound
      WHERE expid = ?
      ",
      params = list(expid)
    )

    if (nrow(existing) > 0) {

      key_existing <- paste(
        existing$nodename,
        round(existing$retention_time, 3),
        round(existing$mass_measured, 6)
      )

      key_new <- paste(
        new_df$nodename,
        round(new_df$retention_time, 3),
        round(new_df$mass_measured, 6)
      )

      new_df <- new_df[!key_new %in% key_existing, ]
    }

    if (nrow(new_df) > 0) {
      dbAppendTable(FT_con, "ms_compound", new_df)
    }
  })
}


```


```{r}
dbGetQuery(
  FT_con,
  "
  SELECT *
  FROM ms_compound
  WHERE expid = 78
    AND anno IS NOT NULL
  LIMIT 10
  "
)


```


```{r}
dbGetQuery(
  FT_con,
  "SELECT COUNT(*) AS n_missing FROM ms_compound WHERE compound_id IS NULL"
)

```

In case of adding new features that were not already in the `ms_compound` table
we create new compound ids for them based on the maximum `compoun_id` in the 
database.
```{r}
dbExecute(
  FT_con,
  "
  CREATE TEMP TABLE tmp_compound_fix AS
  SELECT
    rowid AS rid,
    (
      SELECT COALESCE(MAX(CAST(compound_id AS INTEGER)), 0)
      FROM ms_compound
      WHERE compound_id IS NOT NULL
    )
    + ROW_NUMBER() OVER (ORDER BY rowid) AS new_id
  FROM ms_compound
  WHERE compound_id IS NULL
  "
)

dbExecute(
  FT_con,
  "CREATE INDEX idx_tmp_compound_fix_rid ON tmp_compound_fix(rid)"
)

dbExecute(FT_con, "BEGIN TRANSACTION")

```
we incremented the compound_ids first using a temporary table because the 
`compound_id` column is defined as "TEXT", then we copy back the new ids to the 
`ms_compound` table.
```{r}
dbExecute(
  FT_con,
  "
  UPDATE ms_compound
  SET compound_id = (
    SELECT new_id
    FROM tmp_compound_fix
    WHERE tmp_compound_fix.rid = ms_compound.rowid
  )
  WHERE compound_id IS NULL
  "
)

dbExecute(FT_con, "COMMIT")


```


```{r}
dbExecute(FT_con, "DROP TABLE tmp_compound_fix")

```


```{r}
dbGetQuery(
  FT_con,
  "SELECT COUNT(*) AS n FROM ms_compound WHERE compound_id IS NULL"
)

```

```{r}
dbGetQuery(
  FT_con,
  "
  SELECT compound_id
  FROM ms_compound
  WHERE expid = 78
    AND anno IS NOT NULL
  LIMIT 10
  "
)

```




```{r}
dbGetQuery(
  FT_con,
  "
  SELECT compound_id, nodename, expid
  FROM ms_compound
  ORDER BY compound_id DESC
  LIMIT 10
  "
)

```
```{r}
dbGetQuery(
  FT_con,
  "
  SELECT *
  FROM ms_compound
  WHERE expid = 78
    AND anno IS NOT NULL
  LIMIT 10
  "
)
```


##The same for QTOF Dynlib + flax negative

Dynlib Qtof positive doesn't contain the xcms.txt file in any of its experiments 
