---
title: "Add the grouping and annotation columns to the sql database"
format: html
server: shiny
author: "Ahlam Mentag"
---

## Shiny Documents

In this quarto document we add the grouping and annotation columns to the new 
sql format of Dynlib FTMS and QTOF negative databases. 


```{r}
library(DBI)
library(RSQLite)
library(xcms)
pth <- getwd()
setwd(pth)
```

Create sql connections to the FTMS and QTOF negative databases
```{r}
FT_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/data/DYL_FLAX_FTMSneg.sqlite"
QTOF_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/data/DYL_FLAX_QTOFneg.sqlite"
assoc_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/New_Dynlib/data/combined_assoc.txt"
FT_con <- DBI::dbConnect(RSQLite::SQLite(), FT_path)
QTOF_con <- DBI::dbConnect(RSQLite::SQLite(), QTOF_path)
```

List tables of FTMS negative data
```{r}
dbListTables(FT_con)
```

List columns names of the ms_compound table
```{r}
dbListFields(FT_con, "ms_compound")
```


```{r}
dbGetQuery(FT_con, "Select * from ms_compound Limit 3 ")
```

Scan all experiments to detect required columns
```{r}
base_dir <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/inst/database_FTMS_neg/Experimenten"

xcms_files <- list.files(
  base_dir,
  pattern = "xcms2.txt$",
  recursive = TRUE,
  full.names = TRUE
)

grp_cols_all <- character()
anno_cols_all <- character()

for (f in xcms_files) {
  x <- read.delim(f, nrows = 1, check.names = FALSE)

  grp_cols_all  <- union(grp_cols_all,
                          grep("^nm\\.Grpo\\.Grp", names(x), value = TRUE))
  anno_cols_all <- union(anno_cols_all,
                          grep("^anno", names(x), value = TRUE))
}

grp_cols_all
anno_cols_all


```

```{r}
extra_cols <- c("isf", "MeAb", "CorrPkGrp", "massMS1", "Comb",
                "con", "CON.new", "subgrp", "CON.new.1")

existing_cols <- dbListFields(FT_con, "ms_compound")

add_if_missing <- function(colname) {
  if (!colname %in% existing_cols) {
    message("Adding column: ", colname)
    dbExecute(
      FT_con,
      sprintf("ALTER TABLE ms_compound ADD COLUMN `%s` TEXT", colname)
    )
  }
}

invisible(lapply(extra_cols, add_if_missing))


```


```{r}
base_dir <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/inst/database_FTMS_neg/Experimenten"
folders <- list.dirs(base_dir, recursive = FALSE)

```

```{r}
grp_cols_all  <- character()
anno_cols_all <- character()

for (f in xcms_files) {
  x <- read.delim(f, nrows = 1, check.names = FALSE)

  grp_cols_all  <- union(
    grp_cols_all,
    grep("^nm\\.Grpo\\.Grp", names(x), value = TRUE)
  )

  anno_cols_all <- union(
    anno_cols_all,
    grep("^anno", names(x), value = TRUE)
  )
}

grp_cols_all
anno_cols_all


```

```{r}
dbExecute(FT_con, "
CREATE INDEX IF NOT EXISTS idx_ms_compound_lookup
ON ms_compound (expid, nodename, retention_time, mass_measured)
")

```

Extracting the experiment number from each experiment folder in the old dynlib
database

```{r}
extract_expid <- function(folder_name) {
  as.integer(sub(".*Exp([0-9]+).*", "\\1", folder_name))
}

```

```{r}
build_ms_compound_df <- function(xcms, expid, grp_cols_all = NULL, anno_cols_all = NULL, extra_cols = NULL) {

  df <- data.frame(
    nodename       = xcms$name,
    mass_measured  = xcms$mzmed,
    retention_time = xcms$rtmed,
    expid          = expid,
    stringsAsFactors = FALSE
  )

  # add grouping columns dynamically if present
  if (!is.null(grp_cols_all)) {
    for (cl in grp_cols_all) {
      df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
    }
  }

  # add annotation columns dynamically if present
  if (!is.null(anno_cols_all)) {
    for (cl in anno_cols_all) {
      df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
    }
  }

  # add extra metadata columns dynamically if present
  if (!is.null(extra_cols)) {
    for (cl in extra_cols) {
      df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
    }
  }

  df
}


```


Extracting the grouping and annotation columns from the "xcms2.txt" file of each
experiment:

```{r}
folders <- list.dirs(base_dir, recursive = FALSE)

for (fld in folders) {

  expid <- extract_expid(basename(fld))
  if (is.na(expid)) next

  xcms_file <- file.path(fld, "xcms2.txt")
  if (!file.exists(xcms_file)) next

  message("Processing experiment ", expid)

  xcms <- read.delim(
    xcms_file,
    check.names = FALSE,
    stringsAsFactors = FALSE
  )

# detect extra columns you want to include
extra_cols <- c("isf", "MeAb", "CorrPkGrp", "massMS1", "Comb",
                "con", "CON.new", "subgrp", "CON.new.1")

new_df <- build_ms_compound_df(
  xcms, expid, grp_cols_all, anno_cols_all, extra_cols
)


  dbWithTransaction(FT_con, {

    ##remove rows that already exist
    existing <- dbGetQuery(
      FT_con,
      "
      SELECT nodename, retention_time, mass_measured
      FROM ms_compound
      WHERE expid = ?
      ",
      params = list(expid)
    )

    if (nrow(existing) > 0) {

      key_existing <- paste(
        existing$nodename,
        round(existing$retention_time, 3),
        round(existing$mass_measured, 6)
      )

      key_new <- paste(
        new_df$nodename,
        round(new_df$retention_time, 3),
        round(new_df$mass_measured, 6)
      )

      new_df <- new_df[!key_new %in% key_existing, ]
    }

    if (nrow(new_df) > 0) {
      dbAppendTable(FT_con, "ms_compound", new_df)
    }
  })
}


```


```{r}
dbGetQuery(
  FT_con,
  "
  SELECT *
  FROM ms_compound
  WHERE expid = 254
  LIMIT 10
  "
)

```


```{r}
dbGetQuery(
  FT_con,
  "SELECT COUNT(*) AS n_missing FROM ms_compound WHERE compound_id IS NULL"
)

```

In case of adding new features that were not already in the `ms_compound` table
we create new compound ids for them based on the maximum `compoun_id` in the 
database.
```{r}
dbExecute(
  FT_con,
  "
  CREATE TEMP TABLE tmp_compound_fix AS
  SELECT
    rowid AS rid,
    (
      SELECT COALESCE(MAX(CAST(compound_id AS INTEGER)), 0)
      FROM ms_compound
      WHERE compound_id IS NOT NULL
    )
    + ROW_NUMBER() OVER (ORDER BY rowid) AS new_id
  FROM ms_compound
  WHERE compound_id IS NULL
  "
)

dbExecute(
  FT_con,
  "CREATE INDEX idx_tmp_compound_fix_rid ON tmp_compound_fix(rid)"
)

dbExecute(FT_con, "BEGIN TRANSACTION")

```

we incremented the compound_ids first using a temporary table because the 
`compound_id` column is defined as "TEXT", then we copy back the new ids to the 
`ms_compound` table.
```{r}
dbExecute(
  FT_con,
  "
  UPDATE ms_compound
  SET compound_id = (
    SELECT new_id
    FROM tmp_compound_fix
    WHERE tmp_compound_fix.rid = ms_compound.rowid
  )
  WHERE compound_id IS NULL
  "
)

dbExecute(FT_con, "COMMIT")


```


```{r}
dbExecute(FT_con, "DROP TABLE tmp_compound_fix")

```


```{r}
dbGetQuery(
  FT_con,
  "SELECT COUNT(*) AS n FROM ms_compound WHERE compound_id IS NULL"
)

```


```{r}
dbGetQuery(
  FT_con,
  "
  SELECT compound_id, nodename, expid
  FROM ms_compound
  ORDER BY compound_id DESC
  LIMIT 10
  "
)

```

```{r}
dbGetQuery(
  FT_con,
  "
  SELECT *
  FROM ms_compound
  WHERE expid = 1
  LIMIT 10
  "
)

```


```{r}
dbGetQuery(
  FT_con,
  "
  SELECT compound_id, COUNT(*) AS n
  FROM ms_compound
  GROUP BY compound_id
  HAVING n > 1
  "
)

```

##Adding the same annotation and grouping column from the XCMSExperiment object 
##of ftms flax seed data to the corresponding features in the sql database.

Importing the XCMSgrouping object of ftms negative flax seeds data
```{r}
load("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms_grp.RData")
```

```{r}
featureDefinitions(ftms_grp)
```

Define the columns to add
```{r}
# grouping columns
grp_cols_all <- grep(
  "^nm\\.Grpo\\.Grp",
  names(featureDefinitions(ftms_grp)),
  value = TRUE
)

# annotation columns (including ISF block)
anno_cols_all <- c(
  grep("^anno", names(featureDefinitions(ftms_grp)), value = TRUE),
  "isf", "MeAb", "CorrPkGrp", "massMS1", "Comb",
  "con", "CON.new", "subgrp", "CON.new.1"
)

# columns we will sync
sync_cols <- c(grp_cols_all, anno_cols_all)

```

build ftms_df from featureDefinitions
```{r}
build_ms_compound_from_features <- function(ftms, expid, sync_cols) {

  df <- as.data.frame(featureDefinitions(ftms), stringsAsFactors = FALSE)

  out <- data.frame(
    nodename        = rownames(df),
    expid           = expid,
    mass_measured   = df$mzmed,
    retention_time  = df$rtmed,
    stringsAsFactors = FALSE
  )

  for (cl in sync_cols) {
    out[[cl]] <- if (cl %in% names(df)) df[[cl]] else NA
  }

  out
}

```

```{r}
expid <- 266
ftms_df <- build_ms_compound_from_features(ftms_grp, expid, sync_cols)

```

Ensure DB schema contains needed columns

```{r}
existing_cols <- dbListFields(FT_con, "ms_compound")

for (cl in sync_cols) {
  if (!cl %in% existing_cols) {
    dbExecute(
      FT_con,
      sprintf("ALTER TABLE ms_compound ADD COLUMN `%s` TEXT", cl)
    )
  }
}

```

Create temporary table and read the data 

```{r}
dbExecute(FT_con, "DROP TABLE IF EXISTS tmp_ftms")

dbExecute(FT_con, "
CREATE TEMP TABLE tmp_ftms (
  nodename TEXT,
  expid INTEGER,
  mass_measured REAL,
  retention_time REAL
)
")

# add sync columns dynamically
for (cl in sync_cols) {
  dbExecute(
    FT_con,
    sprintf("ALTER TABLE tmp_ftms ADD COLUMN `%s` TEXT", cl)
  )
}

# write data
dbWriteTable(
  FT_con,
  "tmp_ftms",
  ftms_df,
  append = TRUE
)

```

Create indexes
```{r}
dbExecute(FT_con, "
CREATE INDEX IF NOT EXISTS idx_ms_compound_expid_nodename
ON ms_compound (expid, nodename)
")

dbExecute(FT_con, "
CREATE INDEX IF NOT EXISTS temp.idx_tmp_ftms_expid_nodename
ON tmp_ftms (expid, nodename)
")

```

Update existing rows
```{r}
set_sql <- paste(
  sprintf("`%s` = t.`%s`", sync_cols, sync_cols),
  collapse = ", "
)

sql_update <- sprintf("
UPDATE ms_compound
SET
  mass_measured  = t.mass_measured,
  retention_time = t.retention_time,
  %s
FROM tmp_ftms t
WHERE ms_compound.expid = t.expid
  AND ms_compound.nodename = t.nodename
", set_sql)

dbExecute(FT_con, sql_update)



```


Insert missing rows
```{r}
insert_cols <- c("nodename", "expid", "mass_measured", "retention_time", sync_cols)

sql_insert <- sprintf("
INSERT INTO ms_compound (%s)
SELECT %s
FROM tmp_ftms t
WHERE NOT EXISTS (
  SELECT 1 FROM ms_compound m
  WHERE m.expid = t.expid
    AND m.nodename = t.nodename
)
",
  paste(sprintf("`%s`", insert_cols), collapse = ", "),
  paste(sprintf("t.`%s`", insert_cols), collapse = ", ")
)

dbExecute(FT_con, sql_insert)

```

Verification

```{r}
dbGetQuery(FT_con, "
SELECT nodename, isf, MeAb, CorrPkGrp, massMS1
FROM ms_compound
WHERE expid = 266
  AND isf IS NOT NULL
LIMIT 10
")


```

Now remove the new missing compound_id
```{r}
dbGetQuery(
  FT_con,
  "SELECT COUNT(*) AS n_missing FROM ms_compound WHERE compound_id IS NULL"
)

```


```{r}
dbExecute(
  FT_con,
  "
  CREATE TEMP TABLE tmp_compound_fix AS
  SELECT
    rowid AS rid,
    (
      SELECT COALESCE(MAX(CAST(compound_id AS INTEGER)), 0)
      FROM ms_compound
      WHERE compound_id IS NOT NULL
    )
    + ROW_NUMBER() OVER (ORDER BY rowid) AS new_id
  FROM ms_compound
  WHERE compound_id IS NULL
  "
)

dbExecute(
  FT_con,
  "CREATE INDEX idx_tmp_compound_fix_rid ON tmp_compound_fix(rid)"
)

dbExecute(FT_con, "BEGIN TRANSACTION")

```

we incremented the compound_ids first using a temporary table because the 
`compound_id` column is defined as "TEXT", then we copy back the new ids to the 
`ms_compound` table.
```{r}
dbExecute(
  FT_con,
  "
  UPDATE ms_compound
  SET compound_id = (
    SELECT new_id
    FROM tmp_compound_fix
    WHERE tmp_compound_fix.rid = ms_compound.rowid
  )
  WHERE compound_id IS NULL
  "
)

dbExecute(FT_con, "COMMIT")


```


```{r}
dbExecute(FT_con, "DROP TABLE tmp_compound_fix")

```


```{r}
dbGetQuery(
  FT_con,
  "SELECT COUNT(*) AS n FROM ms_compound WHERE compound_id IS NULL"
)

```


```{r}
dbGetQuery(
  FT_con,
  "
  SELECT compound_id, nodename, expid
  FROM ms_compound
  ORDER BY compound_id DESC
  LIMIT 10
  "
)

```

```{r}
dbGetQuery(
  FT_con,
  "
  SELECT *
  FROM ms_compound
  WHERE expid = 266
  LIMIT 10
  "
)

```

```{r}
dbGetQuery(
  FT_con,
  "
  SELECT compound_id, COUNT(*) AS n
  FROM ms_compound
  GROUP BY compound_id
  HAVING n > 1
  "
)

```

##The same for QTOF Dynlib + flax negative

Scan all experiments to detect required columns
```{r}
base_dir <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/inst/database_QTOF_neg/Experimenten"

xcms_files <- list.files(
  base_dir,
  pattern = "xcms2.txt$",
  recursive = TRUE,
  full.names = TRUE
)

grp_cols_all <- character()
anno_cols_all <- character()

for (f in xcms_files) {
  x <- read.delim(f, nrows = 1, check.names = FALSE)

  grp_cols_all  <- union(grp_cols_all,
                          grep("^nm\\.Grpo\\.Grp", names(x), value = TRUE))
  anno_cols_all <- union(anno_cols_all,
                          grep("^anno", names(x), value = TRUE))
}

grp_cols_all
anno_cols_all


```

```{r}
# Combine all columns you might add
all_cols <- c(grp_cols_all, anno_cols_all, extra_cols)

existing_cols <- dbListFields(QTOF_con, "ms_compound")

add_if_missing <- function(colname) {
  if (!colname %in% existing_cols) {
    message("Adding column: ", colname)
    dbExecute(
      QTOF_con,
      sprintf("ALTER TABLE ms_compound ADD COLUMN `%s` TEXT", colname)
    )
  }
}

# Add all missing columns
invisible(lapply(all_cols, add_if_missing))


```


```{r}
base_dir <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/inst/database_QTOF_neg/Experimenten"
folders <- list.dirs(base_dir, recursive = FALSE)

```

```{r}
grp_cols_all  <- character()
anno_cols_all <- character()

for (f in xcms_files) {
  x <- read.delim(f, nrows = 1, check.names = FALSE)

  grp_cols_all  <- union(
    grp_cols_all,
    grep("^nm\\.Grpo\\.Grp", names(x), value = TRUE)
  )

  anno_cols_all <- union(
    anno_cols_all,
    grep("^anno", names(x), value = TRUE)
  )
}

grp_cols_all
anno_cols_all


```

```{r}
dbExecute(QTOF_con, "
CREATE INDEX IF NOT EXISTS idx_ms_compound_lookup
ON ms_compound (expid, nodename, retention_time, mass_measured)
")

```


Extracting the experiment number from each experiment folder in the old dynlib
database
```{r}
extract_expid <- function(folder_name) {
  as.integer(sub(".*Exp([0-9]+).*", "\\1", folder_name))
}

```

```{r}
build_ms_compound_df <- function(xcms, expid, grp_cols_all = NULL, anno_cols_all = NULL, extra_cols = NULL) {

  df <- data.frame(
    nodename       = xcms$name,
    mass_measured  = xcms$mzmed,
    retention_time = xcms$rtmed,
    expid          = expid,
    stringsAsFactors = FALSE
  )

  # add grouping columns dynamically if present
  if (!is.null(grp_cols_all)) {
    for (cl in grp_cols_all) {
      df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
    }
  }

  # add annotation columns dynamically if present
  if (!is.null(anno_cols_all)) {
    for (cl in anno_cols_all) {
      df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
    }
  }

  # add extra metadata columns dynamically if present
  if (!is.null(extra_cols)) {
    for (cl in extra_cols) {
      df[[cl]] <- if (cl %in% names(xcms)) xcms[[cl]] else NA
    }
  }

  df
}


```


Extracting the grouping and annotation columns from the "xcms2.txt" file of each
experiment:

```{r}
folders <- list.dirs(base_dir, recursive = FALSE)

for (fld in folders) {

  expid <- extract_expid(basename(fld))
  if (is.na(expid)) next

  xcms_file <- file.path(fld, "xcms2.txt")
  if (!file.exists(xcms_file)) next

  message("Processing experiment ", expid)

  xcms <- read.delim(
    xcms_file,
    check.names = FALSE,
    stringsAsFactors = FALSE
  )

# detect extra columns you want to include
extra_cols <- c("isf", "MeAb", "CorrPkGrp", "massMS1", "Comb",
                "con", "CON.new", "subgrp", "CON.new.1")

new_df <- build_ms_compound_df(
  xcms, expid, grp_cols_all, anno_cols_all, extra_cols
)


  dbWithTransaction(QTOF_con, {

    ##remove rows that already exist
    existing <- dbGetQuery(
      QTOF_con,
      "
      SELECT nodename, retention_time, mass_measured
      FROM ms_compound
      WHERE expid = ?
      ",
      params = list(expid)
    )

    if (nrow(existing) > 0) {

      key_existing <- paste(
        existing$nodename,
        round(existing$retention_time, 3),
        round(existing$mass_measured, 6)
      )

      key_new <- paste(
        new_df$nodename,
        round(new_df$retention_time, 3),
        round(new_df$mass_measured, 6)
      )

      new_df <- new_df[!key_new %in% key_existing, ]
    }

    if (nrow(new_df) > 0) {
      dbAppendTable(QTOF_con, "ms_compound", new_df)
    }
  })
}


```


```{r}
dbGetQuery(
  QTOF_con,
  "
  SELECT compound_id, nodename, expid, `nm.Grpo.Grp`, `nm.Grpo.Grp.1`, anno
  FROM ms_compound
  WHERE expid = 44
  LIMIT 10
  "
)

```

```{r}
dbGetQuery(
  QTOF_con,
  "SELECT COUNT(*) AS n_missing FROM ms_compound WHERE compound_id IS NULL"
)

```


```{r}
dbExecute(
  QTOF_con,
  "
  CREATE TEMP TABLE tmp_compound_fix AS
  SELECT
    rowid AS rid,
    (
      SELECT COALESCE(MAX(CAST(compound_id AS INTEGER)), 0)
      FROM ms_compound
      WHERE compound_id IS NOT NULL
    )
    + ROW_NUMBER() OVER (ORDER BY rowid) AS new_id
  FROM ms_compound
  WHERE compound_id IS NULL
  "
)

dbExecute(
  QTOF_con,
  "CREATE INDEX idx_tmp_compound_fix_rid ON tmp_compound_fix(rid)"
)

dbExecute(QTOF_con, "BEGIN TRANSACTION")

```

we incremented the compound_ids first using a temporary table because the 
`compound_id` column is defined as "TEXT", then we copy back the new ids to the 
`ms_compound` table.
```{r}
dbExecute(
  QTOF_con,
  "
  UPDATE ms_compound
  SET compound_id = (
    SELECT new_id
    FROM tmp_compound_fix
    WHERE tmp_compound_fix.rid = ms_compound.rowid
  )
  WHERE compound_id IS NULL
  "
)

dbExecute(QTOF_con, "COMMIT")


```


```{r}
dbExecute(QTOF_con, "DROP TABLE tmp_compound_fix")

```


```{r}
dbGetQuery(
  QTOF_con,
  "SELECT COUNT(*) AS n FROM ms_compound WHERE compound_id IS NULL"
)

```


```{r}
dbGetQuery(
  QTOF_con,
  "
  SELECT compound_id, nodename, expid
  FROM ms_compound
  ORDER BY compound_id DESC
  LIMIT 10
  "
)

```

```{r}
dbGetQuery(
  QTOF_con,
  "
  SELECT *
  FROM ms_compound
  WHERE expid = 44
  LIMIT 10
  "
)

```

```{r}
dbGetQuery(
  QTOF_con,
  "
  SELECT compound_id, COUNT(*) AS n
  FROM ms_compound
  GROUP BY compound_id
  HAVING n > 1
  "
)

```

##Adding the same annotation and grouping column from the XCMSExperiment object 
##of qtof flax seed data to the corresponding features in the sql database.

Importing the XCMSgrouping object of qtof negative flax seeds data
```{r}
load("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/qtof_grp.RData")
```

```{r}
featureDefinitions(ftms_grp)
```

Define the columns to add
```{r}
# grouping columns
grp_cols_all <- grep(
  "^nm\\.Grpo\\.Grp",
  names(featureDefinitions(qtof_grp)),
  value = TRUE
)

# annotation columns (including ISF block)
anno_cols_all <- c(
  grep("^anno", names(featureDefinitions(qtof_grp)), value = TRUE),
  "isf", "MeAb", "CorrPkGrp", "massMS1", "Comb",
  "con", "CON.new", "subgrp", "CON.new.1"
)

# columns we will sync
sync_cols <- c(grp_cols_all, anno_cols_all)

```

build ftms_df from featureDefinitions
```{r}
build_ms_compound_from_features <- function(qtof, expid, sync_cols) {

  df <- as.data.frame(featureDefinitions(qtof), stringsAsFactors = FALSE)

  out <- data.frame(
    nodename        = rownames(df),
    expid           = expid,
    mass_measured   = df$mzmed,
    retention_time  = df$rtmed,
    stringsAsFactors = FALSE
  )

  for (cl in sync_cols) {
    out[[cl]] <- if (cl %in% names(df)) df[[cl]] else NA
  }

  out
}

```

```{r}
expid <- 52 #the experiment number of qtof flax data within the Dynlib database
ftms_df <- build_ms_compound_from_features(qtof_grp, expid, sync_cols)

```

Ensure DB schema contains needed columns

```{r}
existing_cols <- dbListFields(QTOF_con, "ms_compound")

for (cl in sync_cols) {
  if (!cl %in% existing_cols) {
    dbExecute(
      QTOF_con,
      sprintf("ALTER TABLE ms_compound ADD COLUMN `%s` TEXT", cl)
    )
  }
}

```

Create temporary table and read the data 

```{r}
dbExecute(QTOF_con, "DROP TABLE IF EXISTS tmp_ftms")

dbExecute(QTOF_con, "
CREATE TEMP TABLE tmp_ftms (
  nodename TEXT,
  expid INTEGER,
  mass_measured REAL,
  retention_time REAL
)
")

# add sync columns dynamically
for (cl in sync_cols) {
  dbExecute(
    QTOF_con,
    sprintf("ALTER TABLE tmp_ftms ADD COLUMN `%s` TEXT", cl)
  )
}

# write data
dbWriteTable(
  QTOF_con,
  "tmp_ftms",
  ftms_df,
  append = TRUE
)

```

Create indexes
```{r}
dbExecute(QTOF_con, "
CREATE INDEX IF NOT EXISTS idx_ms_compound_expid_nodename
ON ms_compound (expid, nodename)
")

dbExecute(QTOF_con, "
CREATE INDEX IF NOT EXISTS temp.idx_tmp_ftms_expid_nodename
ON tmp_ftms (expid, nodename)
")

```

Update existing rows
```{r}
set_sql <- paste(
  sprintf("`%s` = t.`%s`", sync_cols, sync_cols),
  collapse = ", "
)

sql_update <- sprintf("
UPDATE ms_compound
SET
  mass_measured  = t.mass_measured,
  retention_time = t.retention_time,
  %s
FROM tmp_ftms t
WHERE ms_compound.expid = t.expid
  AND ms_compound.nodename = t.nodename
", set_sql)

dbExecute(QTOF_con, sql_update)



```


Insert missing rows
```{r}
insert_cols <- c("nodename", "expid", "mass_measured", "retention_time", sync_cols)

sql_insert <- sprintf("
INSERT INTO ms_compound (%s)
SELECT %s
FROM tmp_ftms t
WHERE NOT EXISTS (
  SELECT 1 FROM ms_compound m
  WHERE m.expid = t.expid
    AND m.nodename = t.nodename
)
",
  paste(sprintf("`%s`", insert_cols), collapse = ", "),
  paste(sprintf("t.`%s`", insert_cols), collapse = ", ")
)

dbExecute(QTOF_con, sql_insert)

```

Verification

```{r}
dbGetQuery(QTOF_con, "
SELECT nodename, isf, MeAb, CorrPkGrp, massMS1
FROM ms_compound
WHERE expid = 52
  AND isf IS NOT NULL
LIMIT 10
")


```


Now we fill new missing compound_id

```{r}
dbGetQuery(
  QTOF_con,
  "SELECT COUNT(*) AS n_missing FROM ms_compound WHERE compound_id IS NULL"
)

```


```{r}
dbExecute(
  QTOF_con,
  "
  CREATE TEMP TABLE tmp_compound_fix AS
  SELECT
    rowid AS rid,
    (
      SELECT COALESCE(MAX(CAST(compound_id AS INTEGER)), 0)
      FROM ms_compound
      WHERE compound_id IS NOT NULL
    )
    + ROW_NUMBER() OVER (ORDER BY rowid) AS new_id
  FROM ms_compound
  WHERE compound_id IS NULL
  "
)

dbExecute(
  QTOF_con,
  "CREATE INDEX idx_tmp_compound_fix_rid ON tmp_compound_fix(rid)"
)

dbExecute(QTOF_con, "BEGIN TRANSACTION")

```

we incremented the compound_ids first using a temporary table because the 
`compound_id` column is defined as "TEXT", then we copy back the new ids to the 
`ms_compound` table.
```{r}
dbExecute(
  QTOF_con,
  "
  UPDATE ms_compound
  SET compound_id = (
    SELECT new_id
    FROM tmp_compound_fix
    WHERE tmp_compound_fix.rid = ms_compound.rowid
  )
  WHERE compound_id IS NULL
  "
)

dbExecute(QTOF_con, "COMMIT")


```


```{r}
dbExecute(QTOF_con, "DROP TABLE tmp_compound_fix")

```


```{r}
dbGetQuery(
  QTOF_con,
  "SELECT COUNT(*) AS n FROM ms_compound WHERE compound_id IS NULL"
)

```


```{r}
dbGetQuery(
  QTOF_con,
  "
  SELECT compound_id, nodename, expid
  FROM ms_compound
  ORDER BY compound_id DESC
  LIMIT 10
  "
)

```

```{r}
dbGetQuery(
  QTOF_con,
  "
  SELECT *
  FROM ms_compound
  WHERE expid = 52
  LIMIT 10
  "
)

```

```{r}
dbGetQuery(
  QTOF_con,
  "
  SELECT compound_id, COUNT(*) AS n
  FROM ms_compound
  GROUP BY compound_id
  HAVING n > 1
  "
)

```