---
title: "Adapting the DynLib database layout for Spectra"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
---

# Introduction

The DynLib database defines its own database/table layout which prevents the
direct use of the data with R packages from the *RforMassSpectrometry*
initiative. In this document we investigate and the current DynLib database
structure and evaluate whether we can convert and translate the data into the
database layouts used by e.g. the *CompoundDb* or the *MsBackendSql*
packages. These packages define simple and flexible database layouts for storage
of mass spectrometry (MS) data. The data from these databases can be directly
loaded and represented as `Spectra` objects in R, which simplifies the use of
the data in data analysis workflows. Re-using a database layout compatible with
one of the above mentioned packages would simplify any future analysis and usage
of the data since all the infrastructure to handle the MS data is already
provided by the packages from the RforMassSpectrometry initiative.

# Inspect the DynLib database
```{r}
library(CompoundDb)
library(RSQLite)
```


We first inspect the data for one of the databases, the *FTMS positive polarity*
data.

```{r}
pth <- "data/database_FTMS_neg/CSV"
dir(pth)
```

We will next read the various csv files and inspect the data which is stored in
each of them. We first read the *compound* table and list the data type of the
various columns:

```{r}
compound <- read.table(file.path(pth, "compound.csv"), sep = "\t",
                       header = TRUE)
vapply(compound, class, NA_character_)
```

Most of the columns are of type character. Note that `"NULL"` is used for
missing data (instead of `NA`) - we might want to change that.

```{r}
nrow(compound)
length(unique(compound$COMPID))
```

Each row is thus one compound, with it's unique identifier.

```{r}
length(unique(compound$EXPID))
```

The column `EXPID` contains experiment identifiers - and does not contain any
missing data.

We next import the experiment table:

```{r}
experiment <- read.csv(file.path(pth, "experiment.csv"), sep = "\t",
                         header = TRUE)
head(experiment)
```

The data types for the experiment columns:

```{r}
vapply(experiment, class, NA_character_)
```

We also check that the values in `"EXPID"` matches the values in the *compound*
table.

```{r}
all(experiment$EXPID %in% compound$EXPID)
```

That's great.

We next load the *spectra* table. The number of columns and headers
unfortunately does not match:

```{r}
readLines(file.path(pth, "spectra.csv"), n = 4)
```

There are 5 column headers, but only 4 columns with values. We assume that the
last column (`"MS5IDLIST"`) is empty for all spectra and import the remaining 4 columns.

```{r}
spectra <- read.table(file.path(pth, "spectra.csv"), sep = "\t",
                      header = FALSE, skip = 1)
colnames(spectra) <- c("COMPID", "MS2IDLIST", "MS3IDLIST", "MS4IDLIST")
head(spectra)
```

We check the number of missing (`"NULL"`) values per column.

```{r}
nrow(spectra)
vapply(spectra, function(z) sum(z == "NULL"), NA_integer_)
```

We have values for the first two columns while there are no non-missing values
for the MS4 spectra. We next check if all entries in `"COMPID"` are also in the
*compound* table.

```{r}
all(spectra$COMPID %in% compound$COMPID)
all(compound$COMPID %in% spectra$COMPID)
```

OK. Finally also checking if `"COMPID"` and `"MS2IDLIST"` contain the same
information.

```{r}
all(spectra$COMPID == spectra$MS2IDLIST)
```

OK, they are the same.

Next we check the MS2 table.

```{r}
ms2 <- read.table(file.path(pth, "MS2spectra.csv"), sep = "\t", header = TRUE)
head(ms2)
```

We have references to the compound, the precursor m/z (`"PARENT_ION"`) the m/z
and intensity values (`"MS2PEAKLIST"` and `"MS2INTENSITYLIST"`) and also
references to the associated MS3 spectra (`"MS3IDLIST"`). We check again if the
identifiers are consistent with the other tables.

```{r}
all(ms2$COMPID %in% compound$COMPID)
all(compound$COMPID %in% ms2$COMPID)
all(spectra$COMPID %in% ms2$COMPID)
all(ms2$COMPID %in% spectra$COMPID)
```

OK, the `"COMPID"` is consistent between the *compound*, *spectra* and *ms2*
tables. Next we check whether the information in the *spectra* and *ms2* tables
is redundant.

```{r}
all(spectra$COMPID == ms2$COMPID)
all(spectra$MS3IDLIST == ms2$MS3IDLIST)
```

The content is indeed redundant. We next process the table to expand the
`"MS3IDLIST"` and m/z and intensity columns.

```{r}
#' Helper function to replace `"NULL"` with `NA`.
#'
#' @param x `data.frame` or `matrix`.
#'
#' @return `x` with `"NULL"` replaced by `NA`.
fix_na <- function(x) {
    for (i in seq_len(ncol(x))) {
        x[x[, i] == "NULL", i] <- NA
    }
    x
}
ms2 <- fix_na(ms2)

#' expand the MS3IDLIST
ms2$ms3id <- lapply(strsplit(ms2$MS3IDLIST, split = ","), as.integer)

#' expand the m/z values
ms2$mz <- lapply(strsplit(ms2$MS2PEAKLIST, split = ","), as.numeric)
ms2$intensity <- lapply(strsplit(ms2$MS2INTENSITYLIST, split = ","),
                        as.numeric)
```




Next we process the MS3 table.

```{r}
ms3 <- read.table(file.path(pth, "MS3spectra.csv"), sep = "\t", header = TRUE)
head(ms3)

```

The format seems thus to be similar to the MS2 table. We next check again
overlap of IDs with the other tables.

```{r}
all(ms3$COMPID %in% compound$COMPID)
all(compound$COMPID %in% ms3$COMPID)
```

As expected, we don't have an entry for every compound in the *compound*
table. We don't have the reference to the originating MS2 spectrum - but we
should be able to link the MS3 from the MS2 table using the `"MS3IDLIST"`
column.

```{r}
all_ms3_id <- unique(unlist(ms2$ms3id))
all_ms3_id <- all_ms3_id[!is.na(all_ms3_id)]
all(all_ms3_id %in% ms3$MS3ID)
all(ms3$MS3ID %in% all_ms3_id)
```

At last replace the `"NULL"` values and split/extract the m/z and intensity
values.

```{r}
ms3 <- fix_na(ms3)

#' expand the MS4IDLIST
ms3$ms4id <- lapply(strsplit(ms3$MS4IDLIST, split = ","), as.integer)

#' expand the m/z values
ms3$mz <- lapply(strsplit(ms3$MS3PEAKLIST, split = ","), as.numeric)
ms3$intensity <- lapply(strsplit(ms3$MS3INTENSITYLIST, split = ","),
                        as.numeric)
```

Importantly, we need to check if the identifiers within the different tables (in
particular the MS2 and MS3 tables) are **unique** as we want to store these
spectra into the same database table!

```{r}
cmn <- intersect(ms2$MS2ID, ms3$MS3ID)
length(cmn)
```

So, indeed, the two tables share identifiers. Thus we **must** define a new
ID/index for the MS2 and MS3 spectra.

There are no MS4 and MS5 spectra, thus no need to check these files.


# Comments and notes

- `"NULL"` is used for missing data - we should replace that with `NA` to ensure
  the respective columns can be converted properly to the right data type.
- Each compound is associated with one (and only one) MS2 spectrum. The ID of
  the compound and of the MS2 spectrum are always the same.
- Each MS2 spectrum can have no, 1 or multiple MS3 spectra.
- The *spectra* and *MS2spectra* tables contain redundant information. We would
  not necessarily need the *spectra* table.
- Identifiers are consistent between tables, i.e. there are no orphan IDs in any
  table.
- The `"MS2PEAKLIST"` in the MS2 csv file contains the m/z values of the
  fragment peaks. It's integer values for the FTMS because the data was
  generated in the ion trap at unit resolution.


# Restructuring the database to the `CompDb` format

Some notes and steps we will do:

- Replace all `"NULL"` with `NA`.
- All MS spectra should be stored into one database table
- We use the `acquisitionNum` and `precScanNum` spectra variables to reference
  between the MS 2 and MSn spectra. The (integer) value in `precScanNum` of a
  MSn spectrum is always the spectrum ID of the parent MS spectrum and refers to
  its `acquisitionNum`. With this we can use the `filterPrecursorScan()`
  function from *Spectra* to get all related spectra from the database.
- We need to define new spectrum identifiers as the identifiers for MS2 and MS3
  spectra are not unique.

## Compound and experiment data

The compound information should ideally be stored in the *ms_compound* table. We
can either split the experiment information into a separate table, or combine
that into the *ms_compound*. This depends a bit on the amount of redundant data
that would have to be stored into *ms_compound*, i.e. how many compounds to we
generally have per experiment. We will store the compound  and the experiment
tables to the database *as is* and only convert the upper case column names to
lower case and replacing `"COMPID"` to `"compound_id"`. We need also to add
additional required column names to the *ms_compound* database table, but can
fill these with missing values.

We first fix and create the *ms_compound* database table.

```{r}
ms_compound <- compound
colnames(ms_compound) <- tolower(colnames(ms_compound))
#' Rename "COMPID" -> compound_id
colnames(ms_compound) <- sub("^compid", "compound_id", colnames(ms_compound))
ms_compound$compound_id <- as.character(ms_compound$compound_id)
#' Rename compname -> name
colnames(ms_compound) <- sub("^compname", "name", colnames(ms_compound))
#' Rename mass_theoretical -> exactmass
colnames(ms_compound) <- sub("^mass_theoretical", "exactmass",
                             colnames(ms_compound))
#' Replace NULL with NA
ms_compound <- fix_na(ms_compound)
#' We add the remaining required columns
ms_compound$inchi <- NA_character_
ms_compound$inchikey <- NA_character_

#' Ensure columns have the right data type.
ms_compound$exactmass <- as.numeric(ms_compound$exactmass)
ms_compound$ppm_deviation <- as.numeric(ms_compound$ppm_deviation)
```

Next we reformat the *experiment* database table.

```{r}
colnames(experiment) <- tolower(colnames(experiment))
```


## Collapsing the MS spectra data

Ideally, all MS spectra should go into two tables, one with the metadata
(*spectra variables*) and one with the mass peak data (*peaks data*). Also,
spectra from all MS levels should be stored into the same table, keeping however
the link/association to the respective precursor scan. This information (stored
as `"MS2IDLIST"`, `"MS3IDLIST"` etc in DynLib) would need to be added to the
`precScanNum` spectra variable, i.e. the ID (=index) of the spectrum should be
stored in `acquisitionNum` and the parent spectrum in `precScanNum`.

Below we create the respective tables starting with the MS2 spectra. Required
database table columns in this table are:

- `"spectrum_id"`: an arbitrary ID for the spectrum. Has to be an `integer`.
- `"compound_id"`: the ID of the compound to which the spectrum can be
  associated with. Needs to be `character`.
- `"polarity"`: the polarity (as an `integer`, `0` for negative, `1` for
  positive, `NA` for not set).
- `"collision_energy"`: the collision energy.
- `"predicted"`: whether the spectrum was predicted or measured.
- `"splash"`: the SPLASH of the spectrum.
- `"instrument_type"`: the instrument type.
- `"instrument"`: the name of the instrument.
- `"precursor_mz"`: the precursor m/z (as a `numeric`).

We will need to add these if not already present. In addition, we set the
`"dataOrigin"` spectra variable/column to the present project.

```{r}
#' Define the polarity for the present data: 0 for negative, 1 for positive.
pol <- 0L
proj <- "FTMS_neg"

#' Define the spectrum_id for this MS level:
sid <- seq_len(nrow(ms2))

#' Create/extract the spectra metadata; we might need to add information like
#' collision energy etc later, maybe.
ms_2 <- data.frame(spectrum_id = sid,
                   ms_level = rep(2L, nrow(ms2)),
                   polarity = rep(pol, nrow(ms2)),
                   compound_id = as.character(ms2$COMPID),
                   precursor_mz = ms2$PARENT_ION,
                   instrument = rep(experiment$machine[1L], nrow(ms2)),
                   instrument_type = rep(experiment$mstype[1L], nrow(ms2)),
                   acquisitionNum = sid,
                   precScanNum = rep(NA_integer_, nrow(ms2)),
                   collision_energy = rep(NA_character_, nrow(ms2)),
                   predicted = rep(FALSE, nrow(ms2)),
                   splash = rep(NA_character_, nrow(ms2)),
                   dataOrigin = rep(proj, nrow(ms2)),
                   original_id = ms2$MS2ID)

#' Expand the peaks data to one fragment peak per row
msms_spectrum_peak <- data.frame(
    spectrum_id = rep(sid, lengths(ms2$intensity)),
    mz = as.numeric(unlist(ms2$mz, use.names = FALSE)),
    intensity = as.numeric(unlist(ms2$intensity, use.names = FALSE)))
```

**Note** the m/z values are not sorted increasingly in the MS2 spectra - which
we should ensure they are before adding the `"peak_id"` column!

Next we add the MS3 data. What is a bit odd is that we don't have the MS2ID
column in that table. So, we will use the `"MS3IDLIST"` in the MS2 table to
assign the respective IDs/indices to the table. Also, we need to link the
`"precScanNum"` column to the new `"spectrum_id"` we defined for the MS2
spectra.

```{r}
#' Define spectrum_id for this MS level
sid <- seq(sid[length(sid)] + 1L, length.out = nrow(ms3))
#' Create the content for the MS level 3
ms_3 <- data.frame(spectrum_id = sid,
                   ms_level = rep(3L, nrow(ms3)),
                   polarity = rep(pol, nrow(ms3)),
                   compound_id = as.character(ms3$COMPID),
                   precursor_mz = ms3$PARENT_ION,
                   instrument = rep(experiment$machine[1L], nrow(ms3)),
                   instrument_type = rep(experiment$mstype[1L], nrow(ms3)),
                   acquisitionNum = sid,
                   precScanNum = rep(NA_integer_, nrow(ms3)),
                   collision_energy = rep(NA_character_, nrow(ms3)),
                   predicted = rep(FALSE, nrow(ms3)),
                   splash = rep(NA_character_, nrow(ms3)),
                   dataOrigin = rep(proj, nrow(ms3)),
                   original_id = ms3$MS3ID)

#' Need to assign the spectrum_id of the MS2 spectra into the precScanNum
#' of the associated MS3 spectra. The association is through MS3IDLIST
#' get the mapping between ms2 and ms3 IDs.
map_2_3 <- data.frame(ms2 = rep(ms2$MS2ID, lengths(ms2$ms3id)),
                      ms3 = unlist(ms2$ms3id))
#' Get rid of MS2 spectra without MS3 spectra.
map_2_3 <- map_2_3[!is.na(map_2_3$ms3), ]

#' Need to convert this mapping to the new spectrum_id
id_2_3 <- cbind(ms2 = ms_2$spectrum_id[match(map_2_3$ms2, ms2$MS2ID)],
                ms3 = ms_3$spectrum_id[match(map_2_3$ms3, ms3$MS3ID)])

#' Assign the MS2 ID to the "precScanNum" column of the respective MS3 scan
ms_3$precScanNum[match(id_2_3[, 2L], ms_3$spectrum_id)] <- id_2_3[, 1L]

#' Combine with msms_spectrum
msms_spectrum <- rbind(ms_2, ms_3)

#' Process peaks data
tmp <- data.frame(
    spectrum_id = rep(sid, lengths(ms3$intensity)),
    mz = as.numeric(unlist(ms3$mz, use.names = FALSE)),
    intensity = as.numeric(unlist(ms3$intensity, use.names = FALSE)))
#' Append to the *final* peaks table
msms_spectrum_peak <- rbind(msms_spectrum_peak, tmp)
```

At last we have to ensure that mass (fragment) peaks are increasingly
ordered. Also, we add a *peak_id* column to the table.

```{r}
#' Order the peak by m/z per spectrum
msms_spectrum_peak <- msms_spectrum_peak[
    order(msms_spectrum_peak$spectrum_id, msms_spectrum_peak$mz), ]

#' at last adding the peak_id as a running integer
msms_spectrum_peak$peak_id <- seq_len(nrow(msms_spectrum_peak))
```

Validating that the relationship between the MS2 and MS3 scans is still correct:

```{r}
ms2_id <- ms2$MS2ID[2]
ms3_ids <- ms2$ms3id[[2]]

#' Get the MS2 spectrum
a <- msms_spectrum[msms_spectrum$original_id == ms2_id &
                   msms_spectrum$ms_level == 2, ]
#' Get all associated MS3 spectra
b <- msms_spectrum[msms_spectrum$precScanNum == a$spectrum_id &
                   msms_spectrum$ms_level == 3, ]
all.equal(b$original_id, ms3_ids)

## Repeat for another
ms2_id <- ms2$MS2ID[4]
ms3_ids <- ms2$ms3id[[4]]

#' Get the MS2 spectrum
a <- msms_spectrum[msms_spectrum$original_id == ms2_id &
                   msms_spectrum$ms_level == 2, ]
#' Get all associated MS3 spectra
b <- msms_spectrum[msms_spectrum$precScanNum == a$spectrum_id &
                   msms_spectrum$ms_level == 3, ]
all.equal(b$original_id, ms3_ids)

```


## Export the data

All tables are created and we can next export/write these to a SQLite database.

```{r}
dbf <- file.path("data", "DyLib_FTMS_neg_or.sqlite")
if (file.exists(dbf)) file.remove(dbf)
con <- dbConnect(SQLite(), dbf)
```

As a first table we store a *metadata* table to the database. This table should
contain some general info on the version and content of the database.

```{r}
md <- make_metadata(source = "DynLib FTMS neg", url = NA_character_,
                    source_version = "1.0.0", source_date = "2025-05-15")
dbWriteTable(con, name = "metadata", md, row.names = FALSE)
```

Next we write the compound and experiment data. The `CompDb` database layout
supports multiple names per compound. For this we need to add an additional
(empty) database table *synonym*.

```{r}
#' Write empty synonym table
snm <- data.frame(compound_id = integer(), synonym = character())
dbWriteTable(con, name = "synonym", snm, row.names = FALSE)

#' Write compound table
dbWriteTable(con, name = "ms_compound", ms_compound, row.names = FALSE)

#' Write experiment table
dbWriteTable(con, name = "experiment", experiment, row.names = FALSE)
```

Next we write the MS data to the database.

```{r}
dbWriteTable(con, name = "msms_spectrum_peak", msms_spectrum_peak,
             row.names = FALSE)
dbWriteTable(con, name = "msms_spectrum", msms_spectrum, row.names = FALSE)
```

At last we create the indices.

```{r}
#' experiment
tmp <- dbExecute(con, "create index experiment_expid_idx on experiment (expid)")

#' ms_compound
tmp <- dbExecute(
    con, "create index compound_id_idx on ms_compound (compound_id)")
tmp <- dbExecute(
    con, "create index compound_name_idx on ms_compound (name)")
tmp <- dbExecute(
    con, "create index compound_expid_idx on ms_compound (expid)")

#' msms_spectrum
tmp <- dbExecute(
    con, "create index spectrum_id_idx on msms_spectrum (spectrum_id)")
tmp <- dbExecute(
    con, "create index spectrum_cmp_id_idx on msms_spectrum (compound_id)")
tmp <- dbExecute(
    con, "create index msms_psn_idx on msms_spectrum (precScanNum)")
tmp <- dbExecute(
    con, "create index msms_an_idx on msms_spectrum (acquisitionNum)")


#' msms_spectrum_peak
tmp <- dbExecute(
    con, "create index msms_id_idx on msms_spectrum_peak (spectrum_id)")

dbDisconnect(con)
```
	


#Add feature matrix to the sql databases

```{r}
library(data.table)
library(stringr)

base_path <- "data/database_FTMS_neg/Experimenten"
exp_dirs <- list.dirs(base_path, full.names = TRUE, recursive = FALSE)

read_and_process <- function(dir) {
  nodes_file <- file.path(dir, "nodes.txt")
  if (!file.exists(nodes_file)) return(NULL)

  expid <- as.integer(str_extract(basename(dir), "(?<=Exp)\\d+"))

  dt <- fread(nodes_file, sep = "\t", header = TRUE, strip.white = TRUE, check.names = TRUE)

  # Remove columns that are entirely empty / NA
  dt <- dt[, which(colSums(!is.na(dt) & dt != "" & dt != "NULL") > 0), with = FALSE]

  # Standardize the feature name column
  name_col <- grep("name", names(dt), ignore.case = TRUE, value = TRUE)[1]
  setnames(dt, old = name_col, new = "node_name", skip_absent = TRUE)

  # identify where data starts
  node_col_index <- which(names(dt) == "node_name")

  # Candidate columns = all columns to the right
  candidate_cols <- names(dt)[(node_col_index + 1):ncol(dt)]

  # keep only columns that contain at least 1 numeric value
  data_origin_cols <- candidate_cols[
    sapply(candidate_cols, function(col) any(suppressWarnings(!is.na(as.numeric(dt[[col]])))))
  ]

  # to long format
  dt_long <- melt(
    dt,
    id.vars = "node_name",
    measure.vars = data_origin_cols,
    variable.name = "dataOrigin",
    value.name = "value"
  )

  dt_long[, expid := expid]

  return(dt_long)
}

nodes_long <- rbindlist(lapply(exp_dirs, read_and_process), use.names = TRUE, fill = TRUE)

nodes_long



```



```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS feature_matrix (
    node_name TEXT,
    dataOrigin TEXT,
    Value REAL,
    expid INTEGER,
    FOREIGN KEY (expid) REFERENCES experiment(expid)
  );
")
```
```{r}
dbWriteTable(con, "feature_matrix", nodes_long, append = TRUE, row.names = FALSE)
```


```{r}
dbGetQuery(con, "Select * from feature_matrix where expid = 191 LIMIT 10")
```


## Testing the created database

At last we evaluate whether we can use the created database using the
*CompoundDb*/*Spectra* packages.

```{r}
library(Spectra)
library(CompoundDb)

dbf <- file.path("data", "DyLib_FTMS_neg_Or.sqlite")

#' Load the database as a `CompDb` object
cdb <- CompDb(dbf)
cdb
```

Get access to the compounds:

```{r}
compounds(cdb) |> head()
```


```{r}
metadata(cdb) |> head()

```


```{r}
con <- dbConnect(RSQLite::SQLite(), "data/DyLib_FTMS_neg_or.sqlite")
dbListTables(con)

```


```{r}
dbReadTable(con, "ms_compound")

```

```{r}
dbReadTable(con, "msms_spectrum")
```

```{r}
dbReadTable(con, "feature_matrix")
```


```{r}

df <- dbGetQuery(con, "SELECT * FROM msms_spectrum WHERE ms_level = 3")


head(df)

```

```{r}
tb <- dbGetQuery(con, "SELECT * FROM msms_spectrum WHERE acquisitionNum = 4")


head(tb)
```



```{r}
dbReadTable(con, "experiment")
```

```{r}
dbReadTable(con, "synonym")
```




```{r}
dbReadTable(con, "msms_spectrum")
```



```{r}
df <- dbGetQuery(con, "SELECT * FROM msms_spectrum WHERE ms_level = 3")
df
```




Get access to the MS spectra data.

```{r}
s <- Spectra(cdb)
s
```


```{r}
spectraVariables(s)
```

Only, at present we don't have a simple function to extract the experiment data
- we would need to extract that using SQL queries.



# Session information

```{r}
sessionInfo()
```
