---
title: "Produce mgf files from sqlite datasets of flax experiments"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Rebecca Dauwe, Johannes Rainer"
---

## Shiny Documents

In this quarto document we call the sql_to_mgf() function from the "sql_to_mgf.R"
file to create the mgf files from the flax ftms and qtof experiments we have 
with different spectrum types.

```{r}
pth <- getwd()
setwd(pth)
source("sql_to_mgf.R")
```

#Test on flax ftms data
#FTMS neg flax data with "not_assembled" spectrum_type
```{r}
sql_to_mgf(
  sqlite_path   = "C:/Users/ament/Desktop/Github/RDynLib/data/DYL_FLAX_FTMSneg.sqlite",
  expid         = 266,
  spectrum_type = "not_assembled",
  output_file   = "C:/Users/ament/Desktop/Github/RDynLib/data/DYL_FLAX_FTMSneg.mgf"
)
```

#FTMS neg flax data with "assembled" spectrum_type
```{r}
sql_to_mgf(
  sqlite_path   = "C:/Users/ament/Desktop/Github/RDynLib/data/DYL_FLAX_FTMSneg.sqlite",
  expid         = 266,
  spectrum_type = "assembled",
  output_file   = "C:/Users/ament/Desktop/Github/RDynLib/data/DYL_FLAX_FTMSneg_assembled.mgf"
)
```

#Test on flax QTOF data
#QTOF neg flax data with "intersect_single_energy" spectrum_type
```{r}

sql_to_mgf(
  sqlite_path   = "C:/Users/ament/Desktop/Github/RDynLib/data/DYL_FLAX_QTOFneg.sqlite",
  expid         = 52,
  spectrum_type = "intersect_single_energy",
  output_file   = "C:/Users/ament/Desktop/Github/RDynLib/data/FLAX_QTOFneg_one_energ.mgf"
)
```


#QTOF neg flax data with "merged_MSMS_all_energies" spectrum_type
```{r}
sql_to_mgf(
  sqlite_path   = "C:/Users/ament/Desktop/Github/RDynLib/data/DYL_FLAX_QTOFneg.sqlite",
  expid         = 52,
  spectrum_type = "merged_MSMS_all_energies",
  output_file   = "C:/Users/ament/Desktop/Github/RDynLib/data/DYL_FLAX_QTOFneg.mgf"
)
```

