---
title: "Untitled"
format: html
server: shiny
---

## Shiny Documents

This document represents a workflow for adding ftms flax data as a spectra object
to Dynlib ftmsneg sqlite database.

We start by importing libraries 
```{r}
library(Spectra)
library(CompoundDb)
library(RSQLite)
library(S4Vectors) 
```

Next we load the flax ftmsneg spectra object
```{r}
load("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms_one_tree.RData")
```

```{r}
class(ftms_one_tree)
```

```{r}
spectraData(ftms_one_tree)
```

Here we import the Dynlib sql database and we extract the compDb object from it.
```{r}
#this will be the combined database
con <- dbConnect(RSQLite::SQLite(), "C:/Users/amentag/Desktop/these/These/clones/RDynlib/data/DyLib_FTMS_neg_or.sqlite")
dbListTables(con)
#' Load the database as a `CompDb` object
cdb <- CompDb(con)
cdb
```

Next we create an experiment table for flax data in order to add it 
to experiment table of the Dynlib database

```{r}
#create an expid starting from the maximum one in experiment table of the Dynlib
new_exp_id <- max(dbReadTable(con, "experiment")$expid, na.rm = TRUE) + 1

fftmsneg_exp <- data.frame(
  expid = new_exp_id,
  date = format(Sys.Date(), "NA"),         
  user = "NA",                               
  machine = "FTMS",                           
  column = "NA",               
  mstype = "Ftms",                            
  buffera = "NA",
  bufferb = "NA",
  gradient_time = "NA",
  source = "NA",
  mode = "neg",                              
  species = "NA",               
  tissue = "seed",                              
  ce = "NA",                                   
  meta = "NA",                      
  stringsAsFactors = FALSE
)

# Insert this line to experiment table
dbAppendTable(con, "experiment", fftmsneg_exp)

```
We print the experiment table after insertion
```{r}
dbReadTable(con, "experiment")
```

Next we create a compound dataframe for flax compounds, 
then we add it to ms_compound table of the Dynlib database.

```{r}
#Create the nade name 
n <- length(ftms_one_tree)

if(length(spectraNames(ftms_one_tree)) == 0) {
  names_vec <- paste0("spec_", seq_len(n))
} else {
  names_vec <- spectraNames(ftms_one_tree)
}
#Create compounds_df dataframe for flax seed data
compounds_df <- data.frame(
  compound_id = paste0("cmp_", seq_len(n)), 
  nodename = names_vec,
  name = rep(NA_character_, n),
  formula = rep(NA_character_, n),
  mass_measured = precursorMz(ftms_one_tree),
  exactmass = rep(NA_real_, n),
  ppm_deviation = rep(NA_real_, n),
  retention_time = rtime(ftms_one_tree),
  expid = rep(new_exp_id, n),
  subsid = rep(NA_integer_, n),
  conversion = rep(NA_character_, n),
  wavelen = rep(NA_real_, n),
  smiles = rep(NA_character_, n),
  isotope_ratio = rep(NA_real_, n),
  drift_time = rep(NA_real_, n),
  composition = rep(NA_character_, n),
  inchi = rep(NA_character_, n),
  inchikey = rep(NA_character_, n),
  stringsAsFactors = FALSE
)

# Insert with "dbAppendTable()" function
dbAppendTable(con, "ms_compound", compounds_df)

```


```{r}
#retrieve compound_id from ms_compound table based on the new expid
compound_map <- dbGetQuery(con, paste0(
  "SELECT compound_id, nodename FROM ms_compound WHERE expid = ", new_exp_id
))

```


```{r}
# Create node_name
names_vec <- paste0("spec_", seq_len(length(ftms_one_tree)))

# Affecting those names to ftms_one_tree data
spectraNames(ftms_one_tree) <- names_vec

```


```{r}

compound_map_df <- as.data.frame(compound_map)
#Verify that length of ftms values are equal to the number of spectra
length(compound_map_df$compound_id[match(spectraNames(ftms_one_tree), compound_map_df$nodename)])
#It should return 1222

# Add compound_id column
spectraData(ftms_one_tree)$compound_id <- compound_map_df$compound_id[match(spectraNames(ftms_one_tree), compound_map_df$nodename)]


```

```{r}
all(!is.na(ftms_one_tree$compound_id))  # should return TRUE

```

Here we insert spectra of flax ftmsneg to the dynlib CompoundDB object 
```{r}
insertSpectra(cdb, ftms_one_tree)

```


Verification

```{r}
# the combined database
con <- dbConnect(RSQLite::SQLite(), "C:/Users/amentag/Desktop/these/These/clones/RDynlib/data/DyLib_FTMS_neg_or.sqlite")
dbListTables(con)
#' Load the database as a `CompDb` object
cdb <- CompDb(con)
cdb
dbReadTable(con, "experiment")
```

```{r}
dbGetQuery(con, "
  SELECT * FROM ms_compound 
  ORDER BY compound_id DESC 
  LIMIT 10
")
dbGetQuery(con, "
  SELECT * FROM msms_spectrum_peak 
  ORDER BY Peak_id DESC 
  LIMIT 10
")

dbGetQuery(con, "
  SELECT * FROM msms_spectrum 
  ORDER BY spectrum_id DESC 
  LIMIT 10
")

dbGetQuery(con, "
  SELECT * FROM experiment 
  ORDER BY expid DESC 
  LIMIT 10
")

dbGetQuery(con, "
  SELECT * FROM experiment
  WHERE expid = (SELECT MAX(expid) FROM experiment)
")

dbGetQuery(con, "SELECT COUNT(*) AS nb_lignes FROM experiment")

```


```{r}
#Database before 
con0 <- dbConnect(RSQLite::SQLite(), "C:/Users/amentag/Desktop/copy/data/DyLib_FTMS_neg_or.sqlite")
dbListTables(con0)

dbGetQuery(con0, "
  SELECT * FROM ms_compound 
  ORDER BY compound_id DESC 
  LIMIT 10
")
dbGetQuery(con0, "
  SELECT * FROM msms_spectrum_peak 
  ORDER BY Peak_id DESC 
  LIMIT 10
")
dbGetQuery(con0, "
  SELECT * FROM msms_spectrum 
  ORDER BY spectrum_id DESC 
  LIMIT 10
")

dbGetQuery(con0, "
  SELECT * FROM experiment
  WHERE expid = (SELECT MAX(expid) FROM experiment)
")

dbGetQuery(con0, "SELECT COUNT(*) AS nb_lignes FROM experiment")


```

