---
title: "Untitled"
format: html
server: shiny
---

In this quarto document we call the "sim_functions.R" file code to use functions
for similarity calculation between FTMS data and the RDynLib FTMS database.


```{r}
library(Spectra)
library(CompoundDb)
pth <- getwd()
pth
setwd("C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/Similarity/FTMS")
pth <- getwd()
pth
```


First we load the "sim_functions.R" file that contains functions for similarity 
calculation.
```{r}
source("sim_functions_ftmsneg.R")
```

Here we install the new CompoundDb version
```{r}
#BiocManager::install("RforMassSpectrometry/CompoundDb", ref = "jomain", force = TRUE)
```



Next we convert the "Dynlib_FTMS_neg.sqlite" database to a spectra object
```{r}
library(DBI)
library(RSQLite)
dbf <- file.path(pth,"/data", "DyLib_FTMS_neg_or.sqlite")
db <- dbConnect(SQLite(), dbf)
dbExecute(db, "DROP TABLE IF EXISTS feature_matrix ")
experiment_df <- dbGetQuery(db, "SELECT * FROM experiment")

#here we remove "x", "x.1" columns from experiment table 
experiment_df <- experiment_df[, !names(experiment_df) %in% c("x", "x.1")]
dbExecute(db, "DROP TABLE experiment")
dbWriteTable(db, "experiment", experiment_df)

#the columns have been successfully deleted
dbGetQuery(db, "PRAGMA table_info(experiment)")
dbDisconnect(db)
#we convert the sql database to a Compdb object
cdb <- CompDb(dbf)
#we convert the Compdb object to a spectra object
dy_sps <- Spectra(cdb)
```


```{r}
spectraData(dy_sps, columns = c("rtime"))
```



Here we load the FTMS of flax seed database (it's already a spectra object)  
```{r}
load("data/ftms_one_merged_tree.RData")
ft_neg <- ftms_one_merged_tree
spectraData(ft_neg)
```


Next we calculate the similarity score between the dynlib and flax seed FTMS
negative integuments.


```{r}
result <- symmetric_dotproduct_combined(ft_neg, dy_sps, polarity_query = 0, polarity_target = 0, machine_target = "FTMS", threshold = 0.8)
#save the result to a .RData object
save(result, file = "data/match_DyFtms_neg.RData")
load("data/match_DyFtms_neg.RData")
nrow(result)

print(result)
```


the resulting table has `r nrow(result)` rows, so in this comparison there is 
`r nrow(result)` matches



```{r}
View(result[, c("score","matched_peaks_count","target_spectrum_id","spectrumId",
                "msLevel","target_msLevel","rtime","target_rtime")])

```


