---
title: "Untitled"
format: html
server: shiny
---

In this quarto document we call the "sim_functions.R" file code to use functions
for similarity calculation between FTMS data and the RDynLib FTMS database.


```{r}
library(Spectra)
library(CompoundDb)
```


First we load the "sim_functions.R" file that contains functions for similarity 
calculation.
```{r}
source("sim_functions_ftmsneg.R")
```

Here we install the new CompoundDb version
```{r}
#BiocManager::install("RforMassSpectrometry/CompoundDb", ref = "jomain", force = TRUE)
```



Next we convert the "Dynlib_FTMS_neg.sqlite" database to a spectra object
```{r}
library(DBI)
library(RSQLite)
dbf <- file.path("/data", "DyLib_FTMS_neg2.sqlite")
db <- dbConnect(SQLite(), dbf)

experiment_df <- dbGetQuery(db, "SELECT * FROM experiment")

#here we remove "x", "x.1" columns from experiment table 
experiment_df <- experiment_df[, !names(experiment_df) %in% c("x", "x.1")]
dbExecute(db, "DROP TABLE experiment")
dbWriteTable(db, "experiment", experiment_df)

#the columns have been successfully deleted
dbGetQuery(db, "PRAGMA table_info(experiment)")
dbDisconnect(db)
#we convert the sql database to a Compdb object
cdb <- CompDb(dbf)
#we convert the Compdb object to a spectra object
dy_sps <- Spectra(cdb)
```

Here we load the FTMS of flax seed database (it's already a spectra object)  
```{r}
load("data/ftms_one_tree.RData")
ft_neg <- ftms_one_tree
spectraData(ft_neg)
```


Next we calculate the similarity score between the dynlib and flax seed FTMS
negative integuments.


```{r}
source("sim_functions_with_options.R")
result <- symmetric_dotproduct_combined(ft_neg, dy_sps, polarity_query = 0, polarity_target = 0, machine_target = "FTMS", threshold = 0.8)
#save the result to a .RData object
save(result, file = "data/match_results_ftneg2_corr.RData")
load("data/match_results_ftneg2_corr.RData")

print(result)
```


the resulting table has 1765 rows, that's mean we have 974 matches with the 
two databases



```{r}
sp1 <- ft_neg[spectraData(ft_neg)$scanIndex == 2564]
mz(sp1)
intensity(sp1)
precursorMz(sp1)
sp3 <- dy_sps[spectraData(dy_sps)$acquisitionNum == 146494]
mz(sp3)
intensity(sp3)
precursorMz(sp3)
#plotSpectraMirror(sp1, sp3)
```


`

