      ---
title: "Untitled"
format: html
server: shiny
---

## Shiny Documents

In this quarto document we calculate the similarity score of 4 spectra examples from the dynlib ftmsneg and flax seed ftmsneg databases using :

1.  dynlib_symmetric_dotproduct() function

2.  compareSpectra() from the spectra package

3.  NeutralLossesCosine() from the Python's matchms library

```{r}
library(Spectra)
library(Spectra)
library(SpectriPy)
```

```{r}
source("C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/sim_functions_with_options.R")
```

first example 

```{r}
  
  x <- cbind(mz = c(129.456558227539),
             intensity = c(779.431457519531))
  y <- cbind(mz = c(129, 133, 159),
             intensity = c(339.653961, 119.266136, 133.141098))

```

```{r}
  sps1 <- DataFrame(precursorMz = 177.041)
  mz <- list(129.456558227539)
  intensity <- list(779.431457519531)
  sps1$mz <- mz
  sps1$intensity <- intensity
  sps1<- Spectra(sps1)

  sps2 <- DataFrame(precursorMz = 177.04)
  sps2$mz <- list(c(129, 133, 159))
  sps2$intensity <- list(c(339.653961, 119.266136, 133.141098))
  sps2 <- Spectra(sps2)
```

```{r}
res <- c()

# 1. Similarity with dynlib_symmetric_dotproduct
# (score : 0.9934364)
m <- dynlib_map(x, y, 177.041, 177.04) 
res["DynLib"] <- dynlib_symmetric_dotproduct(m$x, m$y)

# 2. Similarity with compareSpectra + ndotproduct
# (score  : 0.9559187 and 1 common)
res["ndotproduct"] <- compareSpectra(
  sps1, sps2,
  ppm = 0,
  MAPFUN = function(sps1, sps2, ...) joinPeaks(sps1, sps2, tolerance = 0.5),
  FUN = function(sps1, sps2, ...) MsCoreUtils::ndotproduct(sps1, sps2, m = 0.6, n = 2),
  matchedPeaksCount = TRUE
)

# 3. Similarity with NeutralLossesCosine
# (score: 0.977711)
param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = TRUE
)
res["neutral losses cosine"] <- compareSpectriPy(sps1, sps2, param = param)

# 4. Similarity with gnps
res["gnps"] <- c(res, compareSpectra(
                  sps1,
                  sps2,
                  ppm = 0,
                  tolerance = 0.5,
                  MAPFUN = joinPeaksGnps,
                  FUN = MsCoreUtils::gnps
              ))

# Results
print(res)


```
```{r}
plotSpectraMirror(sps1, sps2, tolerance = 0.5)
```

second example

```{r}
 #second example(score = 1)
  x <- cbind(mz = c(128.455490112305),
             intensity = c(11648.4765625))
  y <- cbind(mz = c(128),
             intensity = c(741.387939))


```

```{r}
sps1 <- DataFrame(precursorMz = 257.078)
  mz <- list(c(128.455490112305))
  intensity <- list(c(11648.4765625))
  sps1$mz <- mz
  sps1$intensity <- intensity
  sps1<- Spectra(sps1)

  sps2 <- DataFrame(precursorMz = 257.08)
  sps2$mz <- list(c(128))
  sps2$intensity <- list(c(741.387939))
  sps2 <- Spectra(sps2)
  
```


```{r}
res <- c()

# 1. Similarity with dynlib_symmetric_dotproduct
# (score : 1)
m <- dynlib_map(x, y, 257.078, 257.08) 
res["DynLib"] <- dynlib_symmetric_dotproduct(m$x, m$y)

# 2. Similarity with compareSpectra + ndotproduct
# (score  : 1 and 1 common)
res["ndotproduct"] <- compareSpectra(
  sps1, sps2,
  ppm = 0,
  MAPFUN = function(sps1, sps2, ...) joinPeaks(sps1, sps2, tolerance = 0.5),
  FUN = function(sps1, sps2, ...) MsCoreUtils::ndotproduct(sps1, sps2, m = 0.6, n = 2),
  matchedPeaksCount = TRUE
)

# 3. Similarity with NeutralLossesCosine
# (score: 1)
param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = TRUE
)
res["neutral losses cosine"] <- compareSpectriPy(sps1, sps2, param = param)
# 4. Similarity with gnps
res["gnps"] <- c(res, compareSpectra(
                  sps1,
                  sps2,
                  ppm = 0,
                  tolerance = 0.5,
                  MAPFUN = joinPeaksGnps,
                  FUN = MsCoreUtils::gnps
              ))
# Results
print(res)

```

```{r}
plotSpectraMirror(sps1, sps2, tolerance = 0.5)
```


third example
```{r}
x <- cbind(mz = c(159.619970320805, 315.01799205042),
             intensity = c(94.3498382568359 , 4343.77685546875))
y <- cbind(mz = c(315, 433),
             intensity = c(3391.216064, 104.316986))

```


```{r}
  sps1 <- DataFrame(precursorMz = 477.1023)
  mz <- list(c(159.619970320805, 315.01799205042))
  intensity <- list(c(94.3498382568359 , 4343.77685546875))
  sps1$mz <- mz
  sps1$intensity <- intensity
  sps1<- Spectra(sps1)

  sps2 <- DataFrame(precursorMz = 477.1)
  sps2$mz <- list(c(315, 433))
  sps2$intensity <- list(c(3391.216064, 104.316986))
  sps2 <- Spectra(sps2)
```


```{r}
res <- c()

# 1. Similarity with dynlib_symmetric_dotproduct
# (score : 1)
m <- dynlib_map(x, y, 477.1023, 477.1) 
res["DynLib"] <- dynlib_symmetric_dotproduct(m$x, m$y)

# 2. Similarity with compareSpectra + ndotproduct
# (score  : 0.9999986  and 1 common)
res["ndotproduct"] <- compareSpectra(
  sps1, sps2,
  ppm = 0,
  MAPFUN = function(sps1, sps2, ...) joinPeaks(sps1, sps2, tolerance = 0.5),
  FUN = function(sps1, sps2, ...) MsCoreUtils::ndotproduct(sps1, sps2, m = 0.6, n = 2),
  matchedPeaksCount = TRUE
)

# 3. Similarity with NeutralLossesCosine
# (score: 0.9999993 )
param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = TRUE
)
res["neutral losses cosine"] <- compareSpectriPy(sps1, sps2, param = param)

# 4. Similarity with gnps
res["gnps"] <- c(res, compareSpectra(
                  sps1,
                  sps2,
                  ppm = 0,
                  tolerance = 0.5,
                  MAPFUN = joinPeaksGnps,
                  FUN = MsCoreUtils::gnps
              ))
# Results
print(res)

```
```{r}
plotSpectraMirror(sps1, sps2, tolerance = 0.5)
```


fourth example
```{r}
  x <- cbind(mz = c(300.446106481507),
             intensity = c(668.621826171875))
  y <- cbind(mz = c(179, 254, 255, 271, 272, 300, 301, 313, 316, 317, 343),
             intensity = c(150.978485, 105.706268, 507.930054, 433.778381, 293.672119, 5063.506348, 2164.525146, 124.878708, 308.779663, 837.373535, 281.371155))
  m <- dynlib_map(x, y, 463.03, 463.03)
  res <- dynlib_symmetric_dotproduct(m$x, m$y)
```


```{r}
 sps1 <- DataFrame(precursorMz = 463.03)
  mz <- list(c(300.446106481507))
  intensity <- list(c(668.621826171875))
  sps1$mz <- mz
  sps1$intensity <- intensity
  sps1<- Spectra(sps1)

  sps2 <- DataFrame(precursorMz = 463.03)
  sps2$mz <- list(c(179, 254, 255, 271, 272, 300, 301, 313, 316, 317, 343))
  sps2$intensity <- list(c(150.978485, 105.706268, 507.930054, 433.778381, 293.672119, 5063.506348, 2164.525146, 124.878708, 308.779663, 837.373535, 281.371155))
  sps2 <- Spectra(sps2)
```

```{r}
res <- c()

# 1. Similarity with dynlib_symmetric_dotproduct
# (score : 0.9938881)
m <- dynlib_map(x, y, 463.03, 463.03) 
res["DynLib"] <- dynlib_symmetric_dotproduct(m$x, m$y)

# 2. Similarity with compareSpectra + ndotproduct
# (score  : 0.9666572  and 1 common)
res["ndotproduct"] <- compareSpectra(
  sps1, sps2,
  ppm = 0,
  MAPFUN = function(sps1, sps2, ...) joinPeaks(sps1, sps2, tolerance = 0.5),
  FUN = function(sps1, sps2, ...) MsCoreUtils::ndotproduct(sps1, sps2, m = 0.6, n = 2),
  matchedPeaksCount = TRUE
)

# 3. Similarity with NeutralLossesCosine
# (score: 0.9831873 )
param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = TRUE
)
res["neutral losses cosine"] <- compareSpectriPy(sps1, sps2, param = param)

# 4. Similarity with gnps
res["gnps"] <- c(res, compareSpectra(
                  sps1,
                  sps2,
                  ppm = 0,
                  tolerance = 0.5,
                  MAPFUN = joinPeaksGnps,
                  FUN = MsCoreUtils::gnps
              ))
# Results
print(res)

```
```{r}
plotSpectraMirror(sps1, sps2, tolerance = 0.5)
```



As a conclusion all functions return a similar similarity score for those examples


below we test the 3 functions using a different precursorMz for each spectra
object to take into account the neutral losses.

```{r}
#  (score = 0.9946073)
x <- cbind(mz = c(123, 129.456558227539),
           intensity = c(200, 779.431457519531))
y <- cbind(mz = c(129, 133, 159),
           intensity = c(339.653961, 119.266136, 133.141098))
m <- dynlib_map(x, y, 200, 210)
res <- dynlib_symmetric_dotproduct(m$x, m$y)
res
```



```{r}
# example with neutral loss (score = 0.9946073 if neutral loss considered)
  x <- DataFrame(precursorMz = 200)
  mz <- list(c(123, 129.456558227539))
  intensity <- list(c(200, 779.431457519531))
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)

  y <- DataFrame(precursorMz = 210)
  y$mz <- list(c(129, 133, 159))
  y$intensity <- list(c(339.653961, 119.266136, 133.141098))
  y <- Spectra(y)

  param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = FALSE
   )
  #the similarity score is 0.9831873
  res <- compareSpectriPy(x, y, param = param)
res
```
```{r}
plotSpectraMirror(sps1, sps2, tolerance = 0.5)
```


```{r}
  x <- DataFrame(precursorMz = 11)
  mz <- list(c(129.456558227539, 132))
  intensity <- list(c(779.431457519531, 119.266136))
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)

  y <- DataFrame(precursorMz = 12)
  y$mz <- list(c(129, 133, 159))
  y$intensity <- list(c(339.653961, 119.266136, 133.141098))
  y <- Spectra(y)

  param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = FALSE
   )
  #the similarity score is 0.002907737?????
  res <- compareSpectriPy(x, y, param = param)
  res
```
```{r}
plotSpectraMirror(sps1, sps2, tolerance = 0.5)
```


the  "dynlib_symmetric_dotproduct" takes into accont the common peaks + neutral
losses, in contrast "NeutralLossesCosine" considers just the neutral losses, 
That's why we get a low score in the "NeutralLossesCosine" function.


```{r}
  # the similarity score is 0.9553826 and 1 matched peak

  compareSpectra(
  x,
  y,
  ppm = 0,
  tolerance = 0.5,
  m = 0.6,
  n = 2,
  matchedPeaksCount = TRUE
)

```

Using GNPS similarity...

```{r}
compareSpectra(
    x,
    y,
    ppm = 0,
    tolerance = 0.5,
    MAPFUN = joinPeaksGnps,
    FUN = MsCoreUtils::gnps
)

compareSpectra(
    x,
    y,
    ppm = 0,
    tolerance = 0.5
)

```


## Example with fragments with large differences in intensities

Example with two matching peaks, with differences in intensities. We compare the
DynLib cosine similarity score (which considers also neutral losses) with the
*standard* normalized dot product, the GNPS similarity score (that also
considers neutral loss peaks) and the neutral losses cosine similarity score
from *matchms*.

```{r}
d <- DataFrame(precursorMz = 10.0)
d$mz <- list(c(129.456558227539, 132, 201.1))
d$intensity <- list(c(779.431457519531, 119.266136, 983.3))
x <- Spectra(d)

d <- DataFrame(precursorMz = 0.1)
d$mz <- list(c(129, 133, 159, 201.4))
d$intensity <- list(c(339.653961, 119.266136, 133.141098, 19.3))
y <- Spectra(d)

res <- c()
res <- c(res, compareSpectra(
                  x,
                  y,
                  ppm = 0,
                  tolerance = 0.5,
                  n = 3,
                  m = 0.6,
                  FUN = dynlib_symmetric_dotproduct,
                  MAPFUN = dynlib_map
              ))

res <- c(res, compareSpectra(
                  x,
                  y,
                  ppm = 0,
                  tolerance = 0.5,
                  n = 3,
                  m = 0.6
              ))

#' Note: GNPS ignores n, m
res <- c(res, compareSpectra(
                  x,
                  y,
                  ppm = 0,
                  tolerance = 0.5,
                  MAPFUN = joinPeaksGnps,
                  FUN = MsCoreUtils::gnps
              ))

p <- NeutralLossesCosine(tolerance = 0.5,
                         mz_power = 0.6,
                         intensity_power = 3,
                         ignore_peaks_above_precursor = FALSE)
res <- c(res, compareSpectriPy(x, y, p))
names(res) <- c("DynLib", "dotproduct", "GNPS", "neutral losses cosine")
res
```

```{r, results = "asis"}
library(pander)
tmp <- as.data.frame(res)
colnames(tmp) <- "score"
pandoc.table(tmp, style = "rmarkdown")
```

The DynLib and normalized dot product yield almost identical results while the
GNPS score and neutral losses cosine result in different scores. The difference
for the GNPS implementation is most likely the difference in the `m` and `n`
parameters, while it's unclear why the neutral losses cosine has a score of 0.

As a reference, the mirror plot for the two spectra is shown below:

```{r}
plotSpectraMirror(x, y, tolerance = 0.5)
```

We below repeat the calculations changing the parameters `m` and `n` to evaluate
their impact in the score.

```{r}
res <- c()
res <- c(res, compareSpectra(
                  x,
                  y,
                  ppm = 0,
                  tolerance = 0.5,
                  n = 0.5,
                  m = 0,
                  FUN = dynlib_symmetric_dotproduct,
                  MAPFUN = dynlib_map
              ))

res <- c(res, compareSpectra(
                  x,
                  y,
                  ppm = 0,
                  tolerance = 0.5,
                  n = 0.5,
                  m = 0
              ))

#' Note: GNPS ignores n, m
res <- c(res, compareSpectra(
                  x,
                  y,
                  ppm = 0,
                  tolerance = 0.5,
                  MAPFUN = joinPeaksGnps,
                  FUN = MsCoreUtils::gnps
              ))

p <- NeutralLossesCosine(tolerance = 0.5,
                         mz_power = 0,
                         intensity_power = 0.5,
                         ignore_peaks_above_precursor = FALSE)
res <- c(res, compareSpectriPy(x, y, p))
names(res) <- c("DynLib", "dotproduct", "GNPS", "neutral losses cosine")
res
```

```{r, results = "asis"}
tmp <- as.data.frame(res)
colnames(tmp) <- "score"
pandoc.table(tmp, style = "rmarkdown")
```

Again, the DynLib implementation of the cosine similarity results in highly
similar scores to the normalized dot product.

We next change the precursor m/z of one of the two spectra to simulate neutral
loss fragments.

```{r}
#' With neutral loss match
y$precursorMz <- 11.0
res <- c()
res <- c(res, compareSpectra(
                  x,
                  y,
                  ppm = 0,
                  tolerance = 0.5,
                  n = 3,
                  m = 0.6,
                  FUN = dynlib_symmetric_dotproduct,
                  MAPFUN = dynlib_map
              ))

res <- c(res, compareSpectra(
                  x,
                  y,
                  ppm = 0,
                  tolerance = 0.5,
                  n = 3,
                  m = 0.6
              ))

#' Note: GNPS ignores n, m
res <- c(res, compareSpectra(
                  x,
                  y,
                  ppm = 0,
                  tolerance = 0.5,
                  MAPFUN = joinPeaksGnps,
                  FUN = MsCoreUtils::gnps
              ))

p <- NeutralLossesCosine(tolerance = 0.5,
                         mz_power = 0.6,
                         intensity_power = 3,
                         ignore_peaks_above_precursor = FALSE)
res <- c(res, compareSpectriPy(x, y, p))
names(res) <- c("DynLib", "dotproduct", "GNPS", "neutral losses cosine")
res
```

```{r, results = "asis"}
tmp <- as.data.frame(res)
colnames(tmp) <- "score"
pandoc.table(tmp, style = "rmarkdown")
```

The presence of a neutral losses affects the GNPS score and we now also get a
non-zero score for the neutral loss cosine. As expected, the score for the
normalized dot product does not change. Surprisingly, the DynLib score does not
change. We reduce the parameter `n` to evaluate whether the result changes for
the DynLib score.

```{r}
compareSpectra(
    x,
    y,
    ppm = 0,
    tolerance = 0.5,
    n = 0.5,
    m = 0.6,
    FUN = dynlib_symmetric_dotproduct,
    MAPFUN = dynlib_map
)
```

As a comparison we use the normalized dot product.

```{r}
compareSpectra(
    x,
    y,
    ppm = 0,
    tolerance = 0.5,
    n = 0.5,
    m = 0.6
)
```

Indeed, the DynLib score is higher, suggesting the neutral loss fragment is
indeed considered. We repeat the calculation changing the precursor m/z to
disable neutral loss matches. The score is then expected to be lower, and
similar to the one from the normalized dot product.

```{r}
y$precursorMz <- 123.4
compareSpectra(
    x,
    y,
    ppm = 0,
    tolerance = 0.5,
    n = 0.5,
    m = 0.6,
    FUN = dynlib_symmetric_dotproduct,
    MAPFUN = dynlib_map
)
```

As expected, the score is lower, and similar to the normalized dot product one.

Thus, it seems the settings of the parameters `m` and `n` in the DynLib score
are responsible for differences in the scores compared to e.g. the GNPS-based
score.

# Session information

```{r}
sessionInfo()
```
