---
title: "RDynLib_alignment"
author: "Rebecca Dauwe"
format: html
tbl-cap-location: bottom
editor: Source
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
# Prevent R Notebook from resetting working directory at end of chunks
knitr::opts_knit$set(root.dir = getwd())

```

## Introduction

In this document, we upload the existing DynLib library and RDynLib package and align DynLib subdatabases

Functions: (1) `Aligning_General()`:

Core function that actually performs the alignment between one FT experiment and one QTOF experiment.

Works at the data frame level, manipulating FT (ft) and QTOF (syn) peaks.

Key operations inside:

-   Select FT and QTOF subdb from finlist.

-   Filter each subdb by expnr.ft / expnr.syn.

-   Sort FT peaks by retention time (`order(ft.exp[[1]])`).

-   Pick a central peak (COMPID) for regression or local shortlist.

Call helper functions:

-   `Make_loc_shortlist()` create shortlist of FT peaks.

-   `Remov_empties()` remove unmatched peaks.

-   `Calign()` perform local alignment between FT and QTOF peaks.


-    Apply ESI-mode adjustments if necessary.

-    Return aligned table LCal of matched peaks (FT - QTOF).

Input:

expnr.ft, expnr.syn (expid numbers)

finlist (list of FT and QTOF spectra)

Optional parameters: err, t.ini, chrg, ESImode

Output:

LCal: a data frame of aligned FT–QTOF peaks.



(2) `Aligning_FT_QTOF()`

wrapper function that manages the full alignment workflow for a specific FT run, including interactive steps.

Handles user input, association tables, filtering, outlier removal, regression, and plotting.

Validate that expnr.ft exists in DynLib.

Load / initialize Assoc table (for negative or positive mode).

Find candidate QTOF experiment(s) corresponding to the FT experiment.

Prompt the user to select the QTOF experiment.

Call Aligning_General to generate LCal (aligned peaks).

Refine matches using:

-   matchFTSyn: filter matches based on minIon.

-   RemoveOutliers: remove peaks outside RT deviation range.

Fit RT regression: RegressionPie_LCalign and PlotPie_LCalign (for QC visualization).

Prompt the user to confirm writing connections to the association table.

If confirmed, call `FillAssocFTnQTOFn`: update Assoc with aligned FT - QTOF links.

Return updated Assoc table.

Input:

expnr.ft, finlist

Optional: err, t.ini, lc.err, rng, startpoint, cutoff, minIon, Assoc

Output:

Assoc table (updated with aligned peaks)


```{r}
#| message: false

#' Load required libraries
library(devtools)

#` Uploading RDynLib: Load all functions of the RDynlib directory
load_all()
```


## Uploading DynLib

```{r}
# Hard-coded base path (adjust as needed)
base.dir <- "D:/RDynLib/inst"
finlist <- Get_All_DynLib_files(base.dir)
```

## Local Retention Time Alignment with `LCalign`

The `LCalign` function performs **local retention time alignment** between FT and QTOF chromatograms.

- It takes `t.ini = 5` neighboring FT peaks (by default) and computes **all possible retention differences** between each of these 5 FT m/z values and all possible isomeric QTOF m/z values.
- For **each combination of 5 FT–QTOF pairs**, the **standard deviation of retention time differences** is calculated.
- The combination that **yields the lowest standard deviation** is selected, and the 5 FT–QTOF pairs are added to a **final list (`res.F`)**.
- The local sequence of 5 FT peaks is then shifted **one peak to the right**, and the procedure repeats.

### Multiple Contributions of FT Peaks

- Each FT peak participates in **five overlapping local sequences** (first as the fifth peak, then fourth, etc.), so each FT peak contributes **five FT–QTOF pairs** to `res.F`.
- If a particular FT peak is consistently associated with the **same QTOF isomer** across these sequences, this provides **strong support** for the correct FT–QTOF association.
- If the same FT peak pairs with different QTOF isomers in different sequences, the subsequent regression may be **slightly biased**.

Example:

- Consider a sequence of 7 neighboring FT peaks where the **fourth peak** has two possible QTOF isomers:
  - The first three FT peaks differ by 0.5 min from their QTOF partners -> the algorithm selects the first isomer for the fourth peak to minimize SD.
  - The last three FT peaks differ by 1 min -> the second QTOF isomer is selected for the fourth FT peak.

This overlapping selection mechanism can be seen as a **weighing effect**:

- Multiple contributions of each FT peak to `res.F` reduce bias compared to using **only one single association**, which could be a false positive.
- Note: the **first and last few FT peaks** cannot contribute five times and therefore have less impact on the regression.

### Handling Single QTOF Isomers and False Positives

- If **only one QTOF isomer** is present for a given FT peak, it will be selected.
- If this association is incorrect, it creates a **false positive**, which will appear prominently in `res.F` but is expected to **deviate randomly** from the final regression.
- At least one QTOF isomer is always present due to the prior application of the `Remov_empties` function.

### Usage Note

The `LCalign` function **closely interacts with `Differ_of_retdif`**, which calculates differences of retention times. Together, these functions ensure that **local alignment** is robust, accounts for overlapping sequences, and reduces bias in the regression of retention times.

## Example: Running `LCalign` on FT/QTOF Data
`expnr.ft` and `expnr.syn` correspond to the FT and QTOF experiment numbers you want to align.
`err` specifies the allowed retention time error window (in minutes).
`t.ini`determines how many neighboring FT peaks are used in each local alignment window.
`LCal` contains the aligned FT–QTOF pairs in a data frame ready for downstream regression or plotting.
```{r}
# Make sure your finlist is already loaded
# e.g., finlist <- Get_All_DynLib_files(base.dir)

# Select example experiment numbers (adjust as needed)
expnr.ft <- 191
expnr.syn <- 38

# Optional parameters
err <- 0.02
t.ini <- 5
ESImode <- NULL
chrg <- NULL

# Run local retention time alignment
LCal <- Aligning_General(expnr.ft, expnr.syn, err = err, t.ini = t.ini,
                         finlist = finlist, ESImode = ESImode, chrg = chrg)

# Inspect first few aligned FT–QTOF pairs
head(LCal)



The function `Aligning_General()` does the actual alignment between two subdatabases.


The function `Aligning_FT_QTOF()` aligns one FT (Orbitrap / FT) experiment (expnr.ft) with a QTOF (Synapt/QTOF) experiment and (optionally) adds connections into an association table.

```{r}
expnr.ft <- 191
expnr.syn <- 38
LCal <- Aligning_General(expnr.ft, expnr.syn, err=0.02, t.ini=5, finlist)
Assoc <- Aligning_FT_QTOF(expnr.ft, finlist)

```


