---
title: "RDynLib_alignment"
author: "Rebecca Dauwe"
format: html
tbl-cap-location: bottom
editor: Source
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
# Prevent R Notebook from resetting working directory at end of chunks
knitr::opts_knit$set(root.dir = getwd())

```

## Introduction

In this document, we upload the existing DynLib library and RDynLib package and align DynLib subdatabases

Functions: (1) `Aligning_General()`:

Core function that actually performs the alignment between one FT experiment and one QTOF experiment.

Works at the data frame level, manipulating FT (ft) and QTOF (syn) peaks.

Key operations inside:

-   Select FT and QTOF subdb from finlist.

-   Filter each subdb by expnr.ft / expnr.syn.

-   Sort FT peaks by retention time (`order(ft.exp[[1]])`).

-   Pick a central peak (COMPID) for regression or local shortlist.

Call helper functions:

-   `Make_loc_shortlist()` create shortlist of FT peaks.

-   `Remov_empties()` remove unmatched peaks.

-   `Calign()` perform local alignment between FT and QTOF peaks.


-    Apply ESI-mode adjustments if necessary.

-    Return aligned table LCal of matched peaks (FT ↔ QTOF).

Input:

expnr.ft, expnr.syn (expid numbers)

finlist (list of FT and QTOF spectra)

Optional parameters: err, t.ini, chrg, ESImode

Output:

LCal: a data frame of aligned FT–QTOF peaks.



(2) `Aligning_FT_QTOF()`

wrapper function that manages the full alignment workflow for a specific FT run, including interactive steps.

Handles user input, association tables, filtering, outlier removal, regression, and plotting.

Validate that expnr.ft exists in DynLib.

Load / initialize Assoc table (for negative or positive mode).

Find candidate QTOF experiment(s) corresponding to the FT experiment.

Prompt the user to select the QTOF experiment.

Call Aligning_General to generate LCal (aligned peaks).

Refine matches using:

-   matchFTSyn: filter matches based on minIon.

-   RemoveOutliers: remove peaks outside RT deviation range.

Fit RT regression: RegressionPie_LCalign and PlotPie_LCalign (for QC visualization).

Prompt the user to confirm writing connections to the association table.

If confirmed, call `FillAssocFTnQTOFn`: update Assoc with aligned FT - QTOF links.

Return updated Assoc table.

Input:

expnr.ft, finlist

Optional: err, t.ini, lc.err, rng, startpoint, cutoff, minIon, Assoc

Output:

Assoc table (updated with aligned peaks)


```{r}
#| message: false

#' Load required libraries
library(devtools)

```
## Uploading DynLib



```{r}
load_all()
base.dir <- getwd()

finlist <- Get_All_DynLib_files(base.dir)
```

The function `Aligning_FT_QTOF()` aligns one FT (Orbitrap / FT) experiment (expnr.ft) with a QTOF (Synapt/QTOF) experiment and (optionally) adds connections into an association table.

```{r}
expnr.ft <- 191
expnr.syn <- 38
LCal <- Aligning_General(expnr.ft, expnr.syn, err=0.02, t.ini=5, finlist)
Assoc <- Aligning_FT_QTOF(expnr.ft, finlist)

```


