---
title: "RDynLib_alignment"
author: "Rebecca Dauwe"
format: html
tbl-cap-location: bottom
editor: source
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
# Prevent R Notebook from resetting working directory at end of chunks
knitr::opts_knit$set(root.dir = getwd())

```

# Introduction

In this document, we upload the existing DynLib library and RDynLib package and align DynLib subdatabases

# Upload the RDynLib package

All functions of the RDynLib package are in the R subfolder. By running `devtools::load_all`we can immediately use and test all functions in the RDynLib folder and subfolders.

```{r}
#| message: false

#' Load required libraries
library(devtools)

#` Uploading RDynLib: Load all functions of the RDynlib directory
load_all()
```

# Upload the DynLib database

The function `Get_All_DynLib_files` uploads the DynLib database to `finlist`. `finlist`is a list list with the following content: 
\[\[1\] "finlist_FTn" 
\[\[2\]\] "finlist_FTp" 
\[\[3\]\] "finlist_Synn" 
\[\[4\]\] "finlist_Synp" 
\[\[5\]\] "Assoc_COMPID" #DynLibDBassociation.txt 
\[\[6\]\] "Assoc_COMPID2" 
\[\[7\]\] "Assoc_COMPID3" 
\[\[8\]\] "Assoc_EXPID" #subDBexperiment.txt 
\[\[9\]\] "MS1feat" 
\[\[10\]\] "neut" 
\[\[11\]\] "frag"

```{r}
# Hard-coded base path (adjust as needed)
base.dir <- "D:/RDynLib/inst"
finlist <- Get_All_DynLib_files(base.dir)
```

# Aligning experiments in FTMS versus QTOF Sub-Databases

An FTMS analysis and a QTOF analysis in the same mode (positive or negative), both uploaded in the DynLib database, can be aligned using the `Aligning_FT_QTOF()` function. The `Aligning_FT_QTOF()` function uses a combination of other functions. Here we will go through the process step by step.

## (1) Run local retention time alignment with `LCalign`

The `LCalign` function performs **local retention time alignment** between FT and QTOF chromatograms.

### Input

The minimal input of `LCalign` is the **COMPIDs**, **m/z values** and **retention time values** of **an ftms experiment** and **a qtof experiment**. 
In the current function: the DynLib database is given in the `finlist` argument and the subDatabases are selected through the `chrg` and `ESImode` arguments:

If `chrg = NULL`, and `ESImode = NULL`, finlist\[\[1\]\] #FTMS negative and finlist\[\[3\]\] #QTOF negative are selected.

If `chrg = "pos"`, and `ESImode = NULL`, finlist\[\[2\]\] #FTMS positive and finlist\[\[4\]\] #QTOF positive are selected.

If `chrg = NULL`, and `ESImode = "opposite"`, finlist\[\[2\]\] #FTMS positive and finlist\[\[3\]\] #QTOF negative are selected.

If `chrg = "pos"`, and `ESImode = "opposite"`, finlist\[\[1\]\] #FTMS negative and finlist\[\[4\]\] #QTOF positive are selected.

The experiment number is given by the `expnr.ft` and `expnr.syn` arguments.

These data are accessed in `finlist`as follows:
finlist[[1]] # FTMS_neg sub-database
finlist[[1]][[1]] # compound.csv
finlist[[1]][[1]][,1] # RETENTION_TIME
finlist[[1]][[1]][,2] # MASS_MEASURED
finlist[[1]][[1]][,3] # COMPID
finlist[[1]][[1]][,4] # EXPID

### Output

`LCal` contains the aligned FT–QTOF pairs in a data frame ready for downstream regression or plotting.

### Arguments

`Àligning_General` is a function with arguments: 
`expnr.ft` and `expnr.syn` correspond to the FT and QTOF experiment numbers you want to align. 
`err` specifies tolerated m/z value error. By default 0.01 in the case of ftms and 0.02 in the case of qtof.
`t.ini` determines how many neighboring FT peaks are used in each local alignment window, by default 5 
`finlist` 
`ESImode = NULL` #if NULL, both experiments are in similar ionization sub-databases 
`chrg = NULL` #if not NULL, an experiment from the QTOFpos sub-database is involved.

-   It takes `t.ini = 5` neighboring FT peaks (by default) and computes **all possible retention differences** between each of these 5 FT m/z values and all possible isomeric QTOF m/z values.
-   For **each combination of 5 FT–QTOF pairs**, the **standard deviation of retention time differences** is calculated.
-   The combination that **yields the lowest standard deviation** is selected, and the 5 FT–QTOF pairs are added to a **final list (`res.F`)**.
-   The local sequence of 5 FT peaks is then shifted **one peak to the right**, and the procedure repeats.

### Multiple Contributions of FT Peaks

-   Each FT peak participates in **five overlapping local sequences** (first as the fifth peak, then fourth, etc.), so each FT peak contributes **five FT–QTOF pairs** to `res.F`.
-   If a particular FT peak is consistently associated with the **same QTOF isomer** across these sequences, this provides **strong support** for the correct FT–QTOF association.
-   If the same FT peak pairs with different QTOF isomers in different sequences, the subsequent regression may be **slightly biased**.

Example:

-   Consider a sequence of 7 neighboring FT peaks where the **fourth peak** has two possible QTOF isomers:
    -   The first three FT peaks differ by 0.5 min from their QTOF partners -\> the algorithm selects the first isomer for the fourth peak to minimize SD.
    -   The last three FT peaks differ by 1 min -\> the second QTOF isomer is selected for the fourth FT peak.

This overlapping selection mechanism can be seen as a **weighing effect**:

-   Multiple contributions of each FT peak to `res.F` reduce bias compared to using **only one single association**, which could be a false positive.
-   Note: the **first and last few FT peaks** cannot contribute five times and therefore have less impact on the regression.

### Handling Single QTOF Isomers and False Positives

-   If **only one QTOF isomer** is present for a given FT peak, it will be selected.
-   If this association is incorrect, it creates a **false positive**, which will appear prominently in `res.F` but is expected to **deviate randomly** from the final regression.
-   At least one QTOF isomer is always present due to the prior application of the `Remov_empties` function.

### Usage Note

The `LCalign` function **closely interacts with `Differ_of_retdif`**, which calculates differences of retention times. Together, these functions ensure that **local alignment** is robust, accounts for overlapping sequences, and reduces bias in the regression of retention times.

### Example: Running `LCalign` on FT/QTOF Data


```{r}
# Make sure your finlist is already loaded
# e.g., finlist <- Get_All_DynLib_files(base.dir)

# Select example experiment numbers
expnr.ft <- 191
expnr.syn <- 38

# Optional parameters
err <- 0.02
t.ini <- 5
ESImode <- NULL
chrg <- NULL

# Run local retention time alignment
LCal <- Aligning_General(expnr.ft, expnr.syn, err = err, t.ini = t.ini,
                         finlist = finlist, ESImode = ESImode, chrg = chrg)

# Inspect first few aligned FT–QTOF pairs
head(LCal)

```

## (2) Remove false positives based on spectral match
The `matchFTSyn` function removes false positive FTm/z, QTOFm/z peak pairs by checking whether a certain percentage of the MS2 spectral peaks can be traced in the MS/MS spectrum. 
By default, 60% of the MS2 peaks should be traced in the MS/MS spectrum as the MS2 spectrum is much less crowded.

### Input

The arguments of `matchFTSyn` are LCal, finlist, minIon=NULL, and chrg=NULL
The the full DynLib database is given in the `finlist` argument and subdatabases are selected through the `chrg` argument, as above.
The MS2 spectra are accessed through 
	 ms2.ft<-finlist[[1]][[3]] #FT neg
	 ms2.syn<-finlist[[3]][[3]] #QTOF neg
	 ms2.ft<-finlist[[2]][[3]] #FT pos
	 ms2.syn<-finlist[[4]][[3]] #QTOF pos

The fraction of FTMS MS2 peaks that should be traced in the QTOF MS/MS spectrum is determined by the `minIon` argument (0.6 by default).
	 
```{r}


LCal <- matchFTSyn(LCal, finlist, minIon = NULL, chrg = NULL)

```

## (3) Remove outliers based on retention time difference
The `RemoveOutliers` function 
looks at the retention time difference between matched peaks (LCal[,6]);
`rng` is the number of standard deviations that the retention time difference can deviate from the mean retention time difference:
computes mean ± (rng × SD);
removes all rows outside that range.

```{r}
LCal <- RemoveOutliers(LCal, rng = 2)
```

## (4) Regression
The `RegressionPie_LCalign` function computes regression between FT and QTOF chromatograms. 
Deviations from a straight curve might occur as different LC systems are involved. Therefore, a piecewise regression is performed including at most two knots. 

By looking at the result from multiple experiments, robust regression
was more accurate than traditional least squares. 
The 'startpoint' allows to decide on the FT retention time from which regression should be started. 

```{r}
rg <- RegressionPie_LCalign(LCal, startpoint = 1)
```

## (5) Plot Regression line
Plot FT retention times (Y-axis; LCal[,2]) in function of Synapt retention times
(X-axis; LCal[,4])

```{r}
PlotPie_LCalign(LCal,rg)
```

# Create Association table

The `Aligning_FT_QTOF()` function is the wrapper function that manages the full above alignment workflow for a specific FT run, including interactive steps.

This function also **uploads the association tables** and **returns an updated Assoc table**:

Before running `Aligning_FT_QTOF()`, **updated versions of the following files must be uploaded**:

-   `subDBexperiment.txt`: Add a new row indicating experiments (expid) from different sub-databases that belong to the same sample.\
-   `DynLibDBassociation.txt` (for FTMS_neg reference) or `DynLibDBassociation2.txt` (for FTMS_pos reference): Add the COMPIDs of the reference subDB experiment.

Upload these files in R using `Get_All_DynLib_files()`:
finlist[[5]] "Assoc_COMPID" #DynLibDBassociation.txt 
finlist[[5]][,1] #COMPIDnegFT
finlist[[5]][,2] #COMPIDposFT 
finlist[[5]][,3] #COMPIDnegSynapt 
finlist[[5]][,4] #COMPIDposSynapt 
finlist[[5]][,5] #COMPIDnegMSE 
finlist[[5]][,6] #COMPIDposMSE

finlist[[6]] "Assoc_COMPID2" #DynLibDBassociation2.txt

finlist[[8]] "Assoc_EXPID" #subDBexperiment.txt 
finlist[[8]][,1] #Expid SYNPOS
finlist[[8]][,2] #Expid SYNNEG 
finlist[[8]][,3] #Expid FTNEG 
finlist[[8]][,4] #Expid FTPOS 

In `Aligning_FT_QTOF()`, these tables are uploaded as:
Assoc<-finlist[[5]]
Assoc_EXPID_FTn<-strsplit(finlist[[8]][,3],",")


`Aligning_FT_QTOF()` returns an updated Assoc table with aligned COMPIDS

### Example: Running `LCalign` on FT/QTOF Data


```{r}
# Make sure your finlist is already loaded
# e.g., finlist <- Get_All_DynLib_files(base.dir)

# Select example experiment numbers
expnr.ft <- 191
expnr.syn <- 38

# Optional parameters
err <- 0.02
t.ini <- 5
ESImode <- NULL
chrg <- NULL

# Run local retention time alignment
Assoc<-Aligning_FT_QTOF(expnr.ft,finlist, )

```

