---
title: "Untitled"
format: html
server: shiny
---

# Introduction

In this document, we upload the existing DynLib library and RDynLib package and align DynLib subdatabases

# Upload the RDynLib package

All functions of the RDynLib package are in the R subfolder. By running `devtools::load_all`we can immediately use and test all functions in the RDynLib folder and subfolders.


```{r}
#| message: false

#' Load required libraries
library(devtools)
#rm(list = c("convert_compid_assoc"))
#` Uploading RDynLib: Load all functions of the RDynlib directory
load_all()
```

```{r}
# Select example experiment numbers
expnr.ft <- 266
expnr.syn <- 52

# Optional parameters
err <- 0.02
t.ini <- 5

#SQL databases paths
FT_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/Similarity/FTMS/data/DYL_FLAX_FTMSneg.sqlite"
QTOF_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/Similarity/QTOF/data/DYL_FLAX_QTOFneg.sqlite"
assoc_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/New_Dynlib/data/combined_assoc.txt"
FT_con <- DBI::dbConnect(RSQLite::SQLite(), FT_path)
QTOF_con <- DBI::dbConnect(RSQLite::SQLite(), QTOF_path)
  
align_flax_SQL <- Aligning_General_SQL(FT_con, QTOF_con, expnr.ft, expnr.syn, err, t.ini)


head(align_flax_SQL)
nrow(align_flax_SQL)
```


## (2) Remove false positives based on spectral match
The `matchFTSyn` function removes false positive FTm/z, QTOFm/z peak pairs by checking whether a certain percentage of the MS2 spectral peaks can be traced in the MS/MS spectrum. 
By default, 60% of the MS2 peaks should be traced in the MS/MS spectrum as the MS2 spectrum is much less crowded.

### Input

```{r}
LCal_sql_flax <- matchFTSyn_SQL(align_flax_SQL, FT_con, QTOF_con, minIon = 0.6,aggregated_Ft = TRUE, aggregated_QTOF = TRUE)
head(LCal_sql_flax)
nrow(LCal_sql_flax)
```

## (3) Remove outliers based on retention time difference
The `RemoveOutliers` function 
looks at the retention time difference between matched peaks (LCal[,6]);
`rng` is the number of standard deviations that the retention time difference can deviate from the mean retention time difference:
computes mean ± (rng × SD);
removes all rows outside that range.

```{r}
LCal_sql_flax <- RemoveOutliers(LCal_sql_flax, rng = 2)
head(LCal_sql_flax)
nrow(LCal_sql_flax)
```

## (4) Regression
The `RegressionPie_LCalign` function computes regression between FT and QTOF chromatograms. 
Deviations from a straight curve might occur as different LC systems are involved. Therefore, a piecewise regression is performed including at most two knots. 

By looking at the result from multiple experiments, robust regression
was more accurate than traditional least squares. 
The 'startpoint' allows to decide on the FT retention time from which regression should be started. 
```{r}
rg_LCal_sql_flax <- RegressionPie_LCalign_SQL(LCal_sql_flax, startpoint = 1)
head(rg_LCal_sql_flax)
```

## (5) Plot Regression line
Plot FT retention times (Y-axis; LCal[,2]) in function of Synapt retention times
(X-axis; LCal[,4])

```{r}
PlotPie_LCalign(LCal_sql_flax,rg_LCal_sql_flax)
```

Calling the global alignment function:

```{r}
Assoc_SQL_flax <-Aligning_FT_QTOF_SQL(
  FT_path = FT_path,
  QTOF_path = QTOF_path,
  FT_expnr =  266,
  QTOF_expnr = 52,
  startpoint = 1,
  save_assoc = FALSE,
   aggregated_Ft = TRUE, 
  aggregated_QTOF = TRUE
)
```

The number of added rows:
```{r}
n_before <- if (exists("Assoc_df")) nrow(Assoc_df) else 0
n_after <- nrow(Assoc_SQL_flax)
n_new <- n_after - n_before
n_new
```

```{r}
Assoc_SQL_flax
```

