---
title: "Untitled"
format: html
server: shiny
---

## Shiny Documents

This Quarto document is made interactive using Shiny. Interactive documents allow readers to modify parameters and see the results immediately. Learn more about Shiny interactive documents at <https://quarto.org/docs/interactive/shiny/>.

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change. This demonstrates how a standard R plot can be made interactive:

```{r}
#| message: false

#' Load required libraries
library(devtools)
library(DBI)
library(RSQLite)
#rm(list = c("convert_compid_assoc"))
#` Uploading RDynLib: Load all functions of the RDynlib directory
load_all()
```

setting the arguments for MSnspecplot_assoc_SQL() function:

```{r}
#directory to the sql databases
sql_dir <- "C:/Users/ament/Desktop/sql_data"

#path to the file with the aligned compounds
assoc_path <- "C:/Users/ament/Desktop/sql_data/assoc_flax.txt"

#compound_id
dbkey <- 101171

#database name
db_name <- "DYL_FLAX_FTMSneg.sqlite"
```

Plot the spectra of a given compound and the corrosponding compounds 
in the "assoc_path".

```{r}
test <- MSnspecplot_assoc_SQL(sql_dir, db_name, dbkey, assoc_path, nr_col = 35, 
                      nr_col2 = 6, lc.err = 0.02, mz.err = 0.001, MS1 = NULL,
                      prcx = 0.6) 
```
The output of 'MSnspecplot_assoc_SQL()' shows that the 45816 is the aligned 
compound from Qtofneg with 101171 compound from FTMSneg.

Plot just the spectra of the 45816 compound from Qtof database:

```{r}
MSnplot_SQL("C:/Users/ament/Desktop/sql_data/DYL_FLAX_QTOFneg.sqlite", 45816)
```

Plot just the spectra of the 101171 compound from FTMS database:
```{r}
MSnplot_SQL("C:/Users/ament/Desktop/sql_data/DYL_FLAX_FTMSneg.sqlite", 101171)
```

Test the MS2plot to plot just the MS2 spectrum
```{r} 
MS2plot_SQL("C:/Users/ament/Desktop/sql_data/DYL_FLAX_FTMSneg.sqlite", 101171)

```


test cspp.display function
```{r}
cspp.display_SQL("C:/Users/ament/Desktop/sql_data/DYL_FLAX_FTMSneg.sqlite", 1, nr_col = 35) 
```

test gnps.display function
```{r}
gnps.display_SQL("C:/Users/ament/Desktop/sql_data/DYL_FLAX_QTOFneg.sqlite", 26, nr_col2 = 6) 

```

test MSMSplot_SQL function

```{r}
prdion <- read.table(
  "C:/Users/ament/Desktop/sql_data/product_ion_candid.txt",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)

neutloss <- read.table(
  "C:/Users/ament/Desktop/sql_data/neutral_loss_candid.txt",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)


MSMSplot_SQL("C:/Users/ament/Desktop/sql_data/DYL_FLAX_QTOFneg.sqlite", 26,
             prdion = prdion, neutloss = neutloss, err = NULL, minum = 2, 
             oldpar = NULL, nl = NULL)
```


```{r}
MSMSspecplot_SQL("C:/Users/ament/Desktop/sql_data/DYL_FLAX_QTOFneg.sqlite", 26, 
                 prdion = prdion, neutloss = neutloss, err = NULL, 
                 minum = NULL, nr_col = NULL,   nr_col2 = NULL)
```


```{r}
cspp <- "C:/Users/ament/Desktop/sql_data/cspp_ftneg.txt"
sql_path <- "C:/Users/ament/Desktop/sql_data/DYL_FLAX_QTOFneg.sqlite"
cspp.tot_SQL(sql_path = sql_path, 1, mzerr = 0.015,
                         cspp = cspp, peakwidth = NULL, IntThres  = 5)
```



```{r}

MultExp_CSPP_SQL("C:/Users/ament/Desktop/sql_data/DYL_FLAX_QTOFneg.sqlite", 
                 startexp = 1, stopexp = 5, peakwidth = NULL, mzerr = 0.015,
                             IntThres  = 5, cspp = cspp)
```

GNPS and CSPP Network

```{r}
Overall.net_SQL(sql_path = sql_path, exp.id = 1, dbkey = 26,
                            nr_of_seq = 2,
                            thr1 = 3,
                            thr2 = 0.1,
                            thr3 = 0.2)
```

```{r}
Overall.net_SQL(sql_path = sql_path, exp.id = 1, dbkey = 20,
                            nr_of_seq = 2,
                            thr1 = 3,
                            thr2 = 0.1,
                            thr3 = 0.2)
```

```{r}
net.lst1 <- net.addit_SQL(sql_path = sql_path, exp.id= 1, nettype="CSPP", thr1=0, thr2=0, thr3=0)
edges.net <- net.lst1[[2]]
head(edges.net)
```

```{r}
net.lst1 <- net.addit_SQL(sql_path = sql_path, exp.id = 1, nettype = "CSPP", thr1 = 0, thr2 = 0, thr3 = 0)
net.lst1
```

