---
title: "Grouping and annotation of feature spectra"
format: html
server: shiny
author: "Ahlam Mentag"
---

## Shiny Documents

In this quarto document we perform the grouping and annotation of the flax ftms 
and qtof flax features by calling the XCMSgrouping_SQL() and XCMSgrouping2_SQL()
functions, in order to add the grouping and annotation columns to the  
ms_compound table in the sql database.

Importing libraries
```{r}
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
library(pander)
library(dplyr)
```

```{r}
setwd("C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/R")
#' Load required libraries
library(devtools)
#rm(list = c("convert_compid_assoc"))
#` Uploading RDynLib: Load all functions of the RDynlib directory
load_all()
```

Here we load the xcms preprocessed ftms and qtof flax objects:
```{r}
load("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms.RData")
load("C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib_qtof/data/qtof.RData")
```



```{r}
#' assuming flax_ftms is the result object from xcms preprocessing,
#' add a columns "fc" with the sample grouping as used/defined by XCMSgrouping 
sampleData(ftms)$fc <- c(1,2,3,4,5,1,2,3,4,5)
sampleData(qtof)$fc <- c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6)
```

```{r}
sampleData(qtof)
```

```{r}
sampleData(ftms)
```

```{r}
featureDefinitions(ftms)
```

```{r}
ftms2 <- XCMSgrouping_SQL(ftms, tR = 1, y10 = 1, y20 = 0.8, y11 = 46.0)
```

```{r}
fd <- featureDefinitions(ftms2)
```

```{r}
original_cols <- c("mzmed","mzmin","mzmax","rtmed","rtmin","rtmax",
    "npeaks","S1","S9","peakidx","ms_level") 
new_cols <- setdiff(colnames(fd), original_cols)
new_cols
```

```{r}
summary(fd[, new_cols])
```


```{r}
table(fd$subgrp, useNA = "ifany")
```

ftms subgrp column contains 3 groups

```{r}
sapply(fd[, new_cols], function(x) length(unique(x)))

```


```{r}
head(fd[, new_cols], 10)

```

```{r}
table(fd$anno.1)
```

This dataset contains 69 adduct and 5867 ann

save the final `XCMSExperiment` with the new columns of grouping and annotation
```{r}
save(ftms2, file = "C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms2.RData")
```

```{r}
load("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms2.RData")
```

Using `XCMSgrouping2()` on the ftms2 object to add/update additional information to the 
`MSExperiment` object:

```{r}
ftms_grp <- XCMSgrouping2_SQL(ftms2, adjPeak = 1)
```

Checking the new columns:
```{r}
fdf <- featureDefinitions(ftms_grp)
```

```{r}
table(fdf$isf)
```

```{r}
table(fdf$Comb)
```
there are 7 rows with values in Comb
 
Save the new object:

```{r}
save(ftms_grp, file = "C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms_grp.RData")
```


grouping and annotating qtof features

```{r}
qtof2 <- XCMSgrouping_SQL(qtof, tR = 1, y10 = 1, y20 = 0.8, y11 = 46.0)
```

```{r}
fdq <- featureDefinitions(qtof2)
```


```{r}
original_cols <- c("mzmed","mzmin","mzmax","rtmed","rtmin","rtmax",
    "npeaks", "S1Ra", "S1Rb", "S1Rc", "S9Ra", "S9Rb", "S9Rc", "peakidx", "ms_level") 
new_cols <- setdiff(colnames(fdq), original_cols)
new_cols
```

```{r}
summary(fdq[, new_cols])
```


```{r}
table(fdq$subgrp, useNA = "ifany")
```

ftms subgrp column contains 3 groups

```{r}
sapply(fdq[, new_cols], function(x) length(unique(x)))

```


```{r}
head(fdq[, new_cols], 10)

```


```{r}
table(fdq$anno.1)
```

this data contains 3487 adduct and 24380 in anno.1 column

```{r}
table(fdq$anno.2)
```

this data contains 27864 ann and 3 dimer in anno.2 column

save the final `XCMSExperiment` with the new columns of grouping and annotation
```{r}
save(qtof2, file = "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib_qtof/data/qtof2.RData")
```

```{r}
load("C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib_qtof/data/qtof2.RData")
```


Using `XCMSgrouping2()` on the qtof2 object to add/update additional information to the 
`MSExperiment` object:

```{r}
qtof_grp <- XCMSgrouping2_SQL(qtof2, adjPeak = 1)
```

Checking the new columns:
```{r}
fdq_grp <- featureDefinitions(qtof_grp)
```

```{r}
table(fdq_grp$isf)
```

```{r}
table(fdq_grp$Comb)
```

the Comb column of this data contains 3 row values
Save the new object:

```{r}
save(qtof_grp, file = "C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/qtof_grp.RData")
```

