---
title: "RDynLib_alignment"
author: "Rebecca Dauwe"
format: html
tbl-cap-location: bottom
editor: source
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

# Introduction

In this document, we upload the existing DynLib library and RDynLib package and align DynLib subdatabases

# Upload the RDynLib package

All functions of the RDynLib package are in the R subfolder. By running `devtools::load_all`we can immediately use and test all functions in the RDynLib folder and subfolders.

```{r}
#| message: false

#' Load required libraries
library(devtools)

#` Uploading RDynLib: Load all functions of the RDynlib directory
load_all()
```



### Example: Running `LCalign` on FT/QTOF Data



```{r}
# Select example experiment numbers
expnr.ft <- 191
expnr.syn <- 38

# Optional parameters
err <- 0.02
t.ini <- 5

#SQL databases paths
FT_path <- "C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/FTMS/data/DYL_FLAX_FTMSneg.sqlite"
QTOF_path <- "C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/QTOF/data/DYL_FLAX_QTOFneg.sqlite"

FT_con <- DBI::dbConnect(RSQLite::SQLite(), FT_path)
QTOF_con <- DBI::dbConnect(RSQLite::SQLite(), QTOF_path)
  
align_SQL <- Aligning_General_SQL(FT_con, QTOF_con, expnr.ft, expnr.syn, err, t.ini)

head(align_SQL)
```


```{r}
# Select example experiment numbers
expnr.ft <- 266
expnr.syn <- 52

# Optional parameters
err <- 0.02
t.ini <- 5

#SQL databases paths
FT_path <- "C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/FTMS/data/DYL_FLAX_FTMSneg.sqlite"
QTOF_path <- "C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/QTOF/data/DYL_FLAX_QTOFneg.sqlite"
align_flax_SQL <- Aligning_General_SQL(FT_path, QTOF_path, expnr.ft, expnr.syn, err, t.ini)

head(align_flax_SQL)
```


## (2) Remove false positives based on spectral match
The `matchFTSyn` function removes false positive FTm/z, QTOFm/z peak pairs by checking whether a certain percentage of the MS2 spectral peaks can be traced in the MS/MS spectrum. 
By default, 60% of the MS2 peaks should be traced in the MS/MS spectrum as the MS2 spectrum is much less crowded.

### Input

```{r}
LCal_sql <- matchFTSyn_SQL(align_SQL, FT_con, QTOF_con, minIon = 0.6)
head(LCal_sql)
nrow(LCal_sql)
```

## (3) Remove outliers based on retention time difference
The `RemoveOutliers` function 
looks at the retention time difference between matched peaks (LCal[,6]);
`rng` is the number of standard deviations that the retention time difference can deviate from the mean retention time difference:
computes mean ± (rng × SD);
removes all rows outside that range.

```{r}
#We use the same function as the last version
LCal_sql <- RemoveOutliers(LCal_sql, rng = 2)
head(LCal_sql)
nrow(LCal_sql)
```

## (4) Regression
The `RegressionPie_LCalign` function computes regression between FT and QTOF chromatograms. 
Deviations from a straight curve might occur as different LC systems are involved. Therefore, a piecewise regression is performed including at most two knots. 

By looking at the result from multiple experiments, robust regression
was more accurate than traditional least squares. 
The 'startpoint' allows to decide on the FT retention time from which regression should be started. 

```{r}
rg_sql <- RegressionPie_LCalign(LCal_sql, startpoint = 1)
head(rg_sql)
```

## (5) Plot Regression line
Plot FT retention times (Y-axis; LCal[,2]) in function of Synapt retention times
(X-axis; LCal[,4])

```{r}
PlotPie_LCalign(LCal_sql,rg_sql)
```

# Create Association table

The `Aligning_FT_QTOF()` function is the wrapper function that manages the full above alignment workflow for a specific FT run, including interactive steps.

This function also **uploads the association tables** and **returns an updated Assoc table**:

Before running `Aligning_FT_QTOF()`, **updated versions of the following files must be uploaded**:

-   `subDBexperiment.txt`: Add a new row indicating experiments (expid) from different sub-databases that belong to the same sample.\
-   `DynLibDBassociation.txt` (for FTMS_neg reference) or `DynLibDBassociation2.txt` (for FTMS_pos reference): Add the COMPIDs of the reference subDB experiment.

Upload these files in R using `Get_All_DynLib_files()`:
finlist[[5]] "Assoc_COMPID" #DynLibDBassociation.txt 
finlist[[5]][,1] #COMPIDnegFT
finlist[[5]][,2] #COMPIDposFT 
finlist[[5]][,3] #COMPIDnegSynapt 
finlist[[5]][,4] #COMPIDposSynapt 
finlist[[5]][,5] #COMPIDnegMSE 
finlist[[5]][,6] #COMPIDposMSE

finlist[[6]] "Assoc_COMPID2" #DynLibDBassociation2.txt

finlist[[8]] "Assoc_EXPID" #subDBexperiment.txt 
finlist[[8]][,1] #Expid SYNPOS
finlist[[8]][,2] #Expid SYNNEG 
finlist[[8]][,3] #Expid FTNEG 
finlist[[8]][,4] #Expid FTPOS 

In `Aligning_FT_QTOF()`, these tables are uploaded as:
Assoc<-finlist[[5]]
Assoc_EXPID_FTn<-strsplit(finlist[[8]][,3],",")


`Aligning_FT_QTOF()` returns an updated Assoc table with aligned COMPIDS

### Example: Running `LCalign` on FT/QTOF Data

```{r}
FT_con <- dbConnect(SQLite(), FT_path)
QTOF_con <- dbConnect(SQLite(), QTOF_path)
fill <- FillAssocFTnQTOFn_SQL(FT_con, QTOF_con, assoc_path, expnr.ft, expnr.syn,
                                       cutoff = 1, rg_sql, lc.err = 1, err, minIon = 0.6,
                                       polarity_ft = 0, polarity_qtof = 0)
```


```{r}


expnr.ft <- 191
FT_path <- "C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/FTMS/data/DYL_FLAX_FTMSneg.sqlite"
QTOF_path <- "C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/QTOF/data/DYL_FLAX_QTOFneg.sqlite"

assoc_path <- "C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib/New_Dynlib/data/combined_assoc.txt"

Assoc_SQL <- Aligning_FT_QTOF_SQL(
  FT_path = FT_path,
  QTOF_path = QTOF_path,
  expnr.ft = 191,
  expnr.syn = 38,
  Assoc = assoc_path,  #
  startpoint = 1
)

                              
```







