---
title: "Untitled"
format: html
server: shiny
---

In this quarto document we call the "sim_functions_qtofneg.R" file code to use 
functions for similarity calculation between the flax seed ftms negative MS/MS 
assembled spectrum and the DynLib QTOF negative database.


```{r}
library(Spectra)
library(CompoundDb)
pth <- getwd()
pth
```


First we load the "sim_functions.R" file that contains functions for similarity 
calculation.
```{r}
source("sim_functions_qtofneg.R")
```

Here we install the new CompoundDb version
```{r}
#BiocManager::install("RforMassSpectrometry/CompoundDb", ref = "jomain", force = TRUE)
```



Next we convert the "DyLib_QTOF_neg2.sqlite" database to a spectra object
```{r}
library(DBI)
library(RSQLite)
dbf <- file.path("data", "DyLib_QTOF_neg.sqlite")
db <- dbConnect(SQLite(), dbf)

experiment_df <- dbGetQuery(db, "SELECT * FROM experiment")

#here we remove "x", "x.1" columns from experiment table 
experiment_df <- experiment_df[, !names(experiment_df) %in% c("x", "x.1")]
dbExecute(db, "DROP TABLE experiment")
dbWriteTable(db, "experiment", experiment_df)

#the columns have been successfully deleted
dbGetQuery(db, "PRAGMA table_info(experiment)")
dbDisconnect(db)
#we convert the sql database to a Compdb object
cdb <- CompDb(dbf)
#we convert the Compdb object to a spectra object
dy_sps <- Spectra(cdb)
```

Here we load the flax seed ftms negative MS/MS assembled spectrum
(it's already a spectra object)  
```{r}
load("data/ftms_assembled.RData")
spectraData(ftms_assembled)
```


Next we calculate the similarity score between the dynlib QTOF
negative integuments and flax seed ftms negative MS/MS assembled spectrum.


```{r}
source("sim_functions_qtofneg.R")
result1 <- sim_final_final(ftms_assembled, dy_sps, polarity_query = 0, polarity_target = 0, machine_target = "Synapt", threshold = 0.8)
#save the result to a .RData object
save(result1, file = "data/match_assembled_ftms.RData")
load("data/match_assembled_ftms.RData")

print(result1)
nrow(result1)
```


the resulting table has 12 rows, which means 12 matches



`

