---
title: "Untitled"
format: html
server: shiny
---

## Shiny Documents

This Quarto document is made interactive using Shiny. Interactive documents allow readers to modify parameters and see the results immediately. Learn more about Shiny interactive documents at <https://quarto.org/docs/interactive/shiny/>.

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change. This demonstrates how a standard R plot can be made interactive:

```{r}
library(Spectra)
library(DBI)
library(dplyr)
library(RSQLite)
register(SerialParam())
pth <- getwd()
pth
```

Loading the sql database of flax qtof negative
```{r}
con_flax <- dbConnect(RSQLite::SQLite(), "data/QTOF_flax_neg.sqlite")
```

Loading the sql database of Dynlib qtof negative
```{r}
con_dyL <- dbConnect(RSQLite::SQLite(), "data/DyLib_QTOF_neg.sqlite")
```


```{r}
dbListTables(con_flax)
```
```{r}
dbListTables(con_dyL)
```
```{r}
# Create new database
con_merged <- dbConnect(SQLite(), "data/DYL_FLAX_QTOFneg.sqlite")
```


```{r}
merge_databases <- function(main_db, add_db, output_db) {
  
  #Connect to databases
  con_main <- dbConnect(SQLite(), main_db)
  con_add  <- dbConnect(SQLite(), add_db)
  con_out  <- dbConnect(SQLite(), output_db)
  
  tables_main <- setdiff(dbListTables(con_main), "sqlite_sequence")
  tables_add  <- setdiff(dbListTables(con_add), "sqlite_sequence")
  #intersect the columns of both databases
  common_tables <- intersect(tolower(tables_main), tolower(tables_add))
  
  #Compute ID shift
  compute_shift <- function(con_main, con_add, table, key) {
    if (!(table %in% tolower(tables_main)) || !(table %in% tolower(tables_add))) return(0)
    main_tbl <- dbReadTable(con_main, tables_main[tolower(tables_main) == table])
    add_tbl  <- dbReadTable(con_add, tables_add[tolower(tables_add) == table])
    if (!key %in% names(main_tbl) || !key %in% names(add_tbl)) return(0)
    main_vals <- suppressWarnings(as.integer(main_tbl[[key]]))
    add_vals  <- suppressWarnings(as.integer(add_tbl[[key]]))
    max_main <- if(any(!is.na(main_vals))) max(main_vals, na.rm=TRUE) else 0
    min_add  <- if(any(!is.na(add_vals)))  min(add_vals, na.rm=TRUE) else 0
    return(max_main + 1 - min_add)
  }
  
  exp_shift      <- compute_shift(con_main, con_add, "experiment", "expid")
  compound_shift <- compute_shift(con_main, con_add, "ms_compound", "compound_id")
  spectrum_shift <- compute_shift(con_main, con_add, "msms_spectrum", "spectrum_id")
  peak_shift     <- compute_shift(con_main, con_add, "msms_spectrum_peak", "peak_id")
  
  #ID shifting function
  shift_ids <- function(df) {
    if ("expid" %in% names(df))        df$expid       <- as.integer(df$expid) + exp_shift
    if ("spectrum_id" %in% names(df))  df$spectrum_id <- as.integer(df$spectrum_id) + spectrum_shift
    if ("peak_id" %in% names(df))      df$peak_id     <- as.integer(df$peak_id) + peak_shift
    if ("compound_id" %in% names(df))  df$compound_id <- as.character(as.integer(df$compound_id) + compound_shift)
    if ("subsid" %in% names(df))       df$subsid      <- as.integer(df$subsid)
    return(df)
  }
  
  #Fix column types
  fix_column_types <- function(df, table_name) {
    if (nrow(df) == 0) return(df)
    
    if (table_name == "msms_spectrum") {
      df$spectrum_id <- as.integer(df$spectrum_id)
      df$compound_id <- as.character(df$compound_id)
      df$ms_level <- as.integer(df$ms_level)
      df$polarity <- as.integer(df$polarity)
      df$precursor_mz <- as.numeric(df$precursor_mz)
      df$precursorIntensity <- as.numeric(df$precursorIntensity)
      df$precursorCharge <- as.integer(df$precursorCharge)
      df$collision_energy <- as.character(df$collision_energy)
      df$isolationWindowLowerMz <- as.numeric(df$isolationWindowLowerMz)
      df$isolationWindowTargetMz <- as.numeric(df$isolationWindowTargetMz)
      df$isolationWindowUpperMz <- as.numeric(df$isolationWindowUpperMz)
      df$peaks_count <- as.integer(df$peaks_count)
      df$totIonCurrent <- as.numeric(df$totIonCurrent)
      df$basePeakMZ <- as.numeric(df$basePeakMZ)
      df$basePeakIntensity <- as.numeric(df$basePeakIntensity)
      df$ionisationEnergy <- as.numeric(df$ionisationEnergy)
      df$lowMZ <- as.numeric(df$lowMZ)
      df$highMZ <- as.numeric(df$highMZ)
      df$mergedScan <- as.integer(df$mergedScan)
      df$mergedResultScanNum <- as.integer(df$mergedResultScanNum)
      df$mergedResultStartScanNum <- as.integer(df$mergedResultStartScanNum)
      df$mergedResultEndScanNum <- as.integer(df$mergedResultEndScanNum)
      df$injectionTime <- as.numeric(df$injectionTime)
      df$filterString <- as.character(df$filterString)
      df$spectrumId <- as.integer(df$spectrumId)
      df$ionMobilityDriftTime <- as.numeric(df$ionMobilityDriftTime)
      df$scanWindowLowerLimit <- as.numeric(df$scanWindowLowerLimit)
      df$scanWindowUpperLimit <- as.numeric(df$scanWindowUpperLimit)
      df$electronBeamEnergy <- as.numeric(df$electronBeamEnergy)
      df$originalPrecursorMz <- as.numeric(df$originalPrecursorMz)
      df$precursorPurity <- as.numeric(df$precursorPurity)
      df$chromPeakRT <- as.numeric(df$chromPeakRT)
      df$chromPeakMz <- as.numeric(df$chromPeakMz)
      df$chromPeakId <- as.character(df$chromPeakId)
      df$rtime <- as.numeric(df$rtime)
      df$scanIndex <- as.integer(df$scanIndex)
      df$dataStorage <- as.character(df$dataStorage)
      df$centroided <- as.integer(df$centroided)
      df$smoothed <- as.integer(df$smoothed)
      df$instrument <- as.character(df$instrument)
      df$splash <- as.character(df$splash)
      df$instrument_type <- as.character(df$instrument_type)
      df$acquisitionNum <- as.integer(df$acquisitionNum)
      df$precScanNum <- as.integer(df$precScanNum)
      df$predicted <- as.numeric(df$predicted)
      df$dataOrigin <- as.character(df$dataOrigin)
      df$original_id <- as.character(df$original_id)
      
    } else if (table_name == "ms_compound") {
      df$compound_id <- as.character(df$compound_id)
      df$expid <- as.integer(df$expid)
      df$exactmass <- as.numeric(df$exactmass)
      df$subsid <- as.integer(df$subsid)
      df$retention_time <- as.numeric(df$retention_time)
      df$mass_measured <- as.numeric(df$mass_measured)
      df$wavelen <- as.numeric(df$wavelen)
      df$isotope_ratio <- as.numeric(df$isotope_ratio)
      df$drift_time <- as.numeric(df$drift_time)
    } else if (table_name == "msms_spectrum_peak") {
      df$peak_id <- as.integer(df$peak_id)
      df$spectrum_id <- as.integer(df$spectrum_id)
      df$mz <- as.numeric(df$mz)
      df$intensity <- as.numeric(df$intensity)
    }
    return(df)
  }
  
  #Merge common tables 
  for(tbl_name in common_tables) {
    tbl_main <- dbReadTable(con_main, tables_main[tolower(tables_main) == tbl_name])
    tbl_add  <- dbReadTable(con_add, tables_add[tolower(tables_add) == tbl_name])
    
    if(nrow(tbl_add) > 0) tbl_add <- shift_ids(tbl_add)
    
    all_cols <- union(names(tbl_main), names(tbl_add))
    for(col in setdiff(all_cols, names(tbl_main))) tbl_main[[col]] <- rep(NA, nrow(tbl_main))
    for(col in setdiff(all_cols, names(tbl_add))) tbl_add[[col]] <- rep(NA, nrow(tbl_add))
    
    tbl_main <- tbl_main[, all_cols]
    tbl_add  <- tbl_add[, all_cols]
    
    tbl_main[] <- lapply(tbl_main, as.character)
    tbl_add[]  <- lapply(tbl_add, as.character)
    
    merged_tbl <- bind_rows(tbl_main, tbl_add)
    
    merged_tbl <- fix_column_types(merged_tbl, tbl_name)
    
    dbWriteTable(con_out, tbl_name, merged_tbl, overwrite = TRUE)
    cat("Merged:", tbl_name, "\n")
  }
  
  # Copy unique tables
  for(tbl_name in setdiff(tables_main, tables_add)) {
    tbl <- dbReadTable(con_main, tbl_name)
    tbl <- fix_column_types(tbl, tbl_name)
    dbWriteTable(con_out, tbl_name, tbl, overwrite = TRUE)
  }
  
  for(tbl_name in setdiff(tables_add, tables_main)) {
    tbl <- dbReadTable(con_add, tbl_name)
    if(nrow(tbl) > 0) tbl <- shift_ids(tbl)
    tbl <- fix_column_types(tbl, tbl_name)
    dbWriteTable(con_out, tbl_name, tbl, overwrite = TRUE)
  }
  
  dbDisconnect(con_main)
  dbDisconnect(con_add)
  dbDisconnect(con_out)
  
}




merge_databases(
  main_db   = "data/DyLib_QTOF_neg.sqlite",
  add_db    = "data/QTOF_flax_neg.sqlite",
  output_db = "data/DYL_FLAX_QTOFneg.sqlite"
)




```


```{r}
con_merged <- dbConnect(SQLite(), "data/DYL_FLAX_QTOFneg.sqlite")
dbListTables(con_merged)
```

```{r}
dbGetQuery(con_merged, "PRAGMA table_info(msms_spectrum);")
dbGetQuery(con_merged, "PRAGMA table_info(ms_compound);")
```


#Ensuring the experiment table has been well merged
```{r}
dbReadTable(con_merged, "experiment")
```

The new experiment has 52 as expid

#Ensuring the ms_compound has been well merged
```{r}
dbGetQuery(con_merged, "Select * from ms_compound where expid = 52 LIMIT 10")
```


52 is the expid of the new added experiment
```{r}
dbGetQuery(con_dyL, "
  SELECT * FROM ms_compound
  ORDER BY CAST(compound_id AS INTEGER) DESC
  LIMIT 10
")

```

The highest compound_id is 45807 in the Dynlib database

```{r}
dbGetQuery(con_merged, "Select * from ms_compound where expid = 52 LIMIT 10")
```

the query results show that the compound_id of the added experiment "52" start by 45808 which is correct

#Ensuring the msms_spectrum table has been well merged

```{r}
dbGetQuery(con_dyL, "
  SELECT * FROM msms_spectrum
  ORDER BY CAST(spectrum_id AS INTEGER) DESC
  LIMIT 10
")
```
The highest spectrum_id is 45807 in the Dynlib database


```{r}
dbGetQuery(con_merged, "Select * from msms_spectrum where compound_id = 45808 Limit 10")
```

the query results show that the spectrum_id of the added experiment "52" start by 45808 which is correct


#Ensuring the feature_matrix table has been well merged

```{r}
dbReadTable(con_merged, "feature_matrix")
```

```{r}
dbGetQuery(con_merged, "Select * from feature_matrix where expid = 52 LIMIT 10")
```

#Converting the merged sql connection to a compdb object
```{r}
library(CompoundDb)

cdb <- CompDb(con_merged)
cdb
```
```{r}
compounds(cdb) |> head()
```

```{r}
sp <- Spectra(cdb)
```


```{r}
spectraData(sp, columns = c("msLevel", "compound_id", "nodename", "rtime", "retention_time"))
```

