---
title: "Untitled"
format: html
server: shiny
---

In this quarto document we call the "sim_functions_qtofneg.R" file code to use 
functions for similarity calculation between QTOF data and the RDynLib QTOF 
negative database.


```{r}
library(Spectra)
library(CompoundDb)
```


First we load the "sim_functions.R" file that contains functions for similarity 
calculation.
```{r}
source("C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/QTOF/sim_functions_qtofneg.R")
```

Here we install the new CompoundDb version
```{r}
#BiocManager::install("RforMassSpectrometry/CompoundDb", ref = "jomain", force = TRUE)
```



Next we convert the "DyLib_QTOF_neg2.sqlite" database to a spectra object
```{r}
library(DBI)
library(RSQLite)
dbf <- file.path("C:/Users/amentag/Desktop/these/These/clones/RDynlib/data", "DyLib_QTOF_neg2.sqlite")
db <- dbConnect(SQLite(), dbf)

experiment_df <- dbGetQuery(db, "SELECT * FROM experiment")

#here we remove "x", "x.1" columns from experiment table 
experiment_df <- experiment_df[, !names(experiment_df) %in% c("x", "x.1")]
dbExecute(db, "DROP TABLE experiment")
dbWriteTable(db, "experiment", experiment_df)

#the columns have been successfully deleted
dbGetQuery(db, "PRAGMA table_info(experiment)")
dbDisconnect(db)
#we convert the sql database to a Compdb object
cdb <- CompDb(dbf)
#we convert the Compdb object to a spectra object
dy_sps <- Spectra(cdb)
```

Here we load the QTOF of flax seed database (it's already a spectra object)  
```{r}
load("C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib_qtof/data/qtof_filtered.RData")
qtof_neg <- qtof_filtered
spectraData(qtof_neg)
```


Next we calculate the similarity score between the dynlib and flax seed QTOF
negative integuments.


```{r}
source("C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/QTOF/sim_functions_qtofneg.R")
result <- sim_final_final(qtof_neg, dy_sps, polarity_query = 0, polarity_target = 0, machine_target = "Synapt", threshold = 0.8)
#save the result to a .RData object
save(result, file = "C:/Users/amentag/Desktop/these/These/clones/RDynlib/data/match_results_qtofneg.RData")
load("C:/Users/amentag/Desktop/these/These/clones/RDynlib/data/match_results_qtofneg.RData")

print(result)
```


the resulting table has 20 rows.



`

