---
title: "Untitled"
format: html
server: shiny
---

In this quarto document we call the "sim_functions_qtofneg.R" file code to use 
functions for similarity calculation between QTOF data and the RDynLib QTOF 
negative database.


```{r}
library(Spectra)
library(CompoundDb)
setwd("C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity")
pth <- getwd()
pth
```


First we load the "sim_functions.R" file that contains functions for similarity 
calculation.
```{r}
source("sim_functions_qtof_ftms.R")
```

Here we install the new CompoundDb version
```{r}
#BiocManager::install("RforMassSpectrometry/CompoundDb", ref = "jomain", force = TRUE)
```



```{r}
load("data/ftms_assembled.RData")
ftms_MS2 <- ftms_assembled
spectraData(ftms_MS2)
```


Here we load the QTOF of flax seed database (it's already a spectra object)  
```{r}
load("data/ms2_qtof.RData")
qtof_neg <- ms2_qtof
spectraData(qtof_neg)
```


Next we calculate the similarity score between the dynlib and flax seed QTOF
negative integuments.


```{r}
#source("sim_functions_qtofneg.R")
result <- sim_final_final(qtof_neg, ftms_MS2, polarity_query = 0, polarity_target = 0, threshold = 0.8)
#save the result to a .RData object
save(result, file = "data/match_qtof_ftms_flax.RData")
load("data/match_qtof_ftms_flax.RData")

print(result)
nrow(result)
```


the resulting table has 16 rows, which means 16 matches



`

