---
title: "Untitled"
format: html
server: shiny
---

In this quarto document we call the "sim_functions.R" file code to use functions for similarity calculation between FTMS data and the RDynLib FTMS database.


```{r}
library(Spectra)
library(CompoundDb)
```


First we load the "sim_functions.R" file that contains functions for similarity calculation
```{r}
source("C:/Users/amentag/Desktop/these/These/similarity/sim_functions.R")
```

Next we load the "Dynlib_FTMS_neg.sqlite" database into a spectra object 
```{r}

dbf <- file.path("C:/Users/amentag/Desktop/these/These/clones/RDynlib/data", "DyLib_FTMS_neg.sqlite")

#' Load the database as a `CompDb` object
cdb <- CompDb(dbf)
dy_sps <- Spectra(cdb)
dy_sps
```

Here we load the FTMS of flax seed database (it's already a spectra object)  
```{r}
load("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms_one_tree.RData")
ft_neg <- ftms_one_tree
spectraData(ft_neg)
```

```{r}


results <- compare_with_metaboannotation(ft_neg, dy_sps, threshold = 0.8)


head(results)


```


