---
title: "Untitled"
format: html
server: shiny
---

In this quarto document we call the "sim_functions.R" file code to use functions for similarity calculation between FTMS data and the RDynLib FTMS database.


```{r}
library(Spectra)
library(CompoundDb)
```


First we load the "sim_functions.R" file that contains functions for similarity calculation
```{r}
source("C:/Users/amentag/Desktop/these/These/similarity/sim_functions.R")
```

Test:

```{r}
x <- matrix(c(250, 100), ncol = 2, byrow = TRUE)  # 300 - 250 = 50
y <- matrix(c(250, 120), ncol = 2, byrow = TRUE)  # 300 - 250 = 50

# Test
map <- dynlib_map(x, y, xPrecursorMz = 300, yPrecursorMz = 300)
print(map)

final_sim(map$x, map$y)
```





Next we load the "Dynlib_FTMS_neg.sqlite" database into a spectra object 
```{r}

dbf <- file.path("C:/Users/amentag/Desktop/these/These/clones/RDynlib/data", "DyLib_FTMS_neg.sqlite")

#' Load the database as a `CompDb` object
cdb <- CompDb(dbf)
dy_sps <- Spectra(cdb)
dy_sps
```

Here we load the FTMS of flax seed database (it's already a spectra object)  
```{r}
load("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms_one_tree.RData")
ft_neg <- ftms_one_tree
spectraData(ft_neg)
```


```{r}
result <- final_sim(ft_neg, dy_sps, threshold = 0.8)
write.csv(result, file = "C:/Users/amentag/Desktop/these/These/clones/RDynlib/data/match_results.csv", 
          row.names = FALSE)

print(result)

```





Test :
```{r}





```


```{r}
'''
library(Spectra)
library(S4Vectors)
library(MetaboAnnotation)

# Spectre 1
spectra_data1 <- DataFrame(
  msLevel = 2L,
  polarity = 1L,
  id = "query1",
  name = "ExampleQuery",
  precursorMz = 300,
  mz = list(c(100, 150, 200)),
  intensity = list(c(10, 100, 50))
)
sps1 <- Spectra(spectra_data1)

# Spectre 2
spectra_data2 <- DataFrame(
  msLevel = 2L,
  polarity = 1L,
  id = "target1",
  name = "ExampleTarget",
  precursorMz = 300,
  mz = list(c(100, 150, 210)),  # Note que 100 et 150 sont communs avec sps1
  intensity = list(c(5, 110, 40))
)
sps2 <- Spectra(spectra_data2)


result <- compare_with_metaboannotation(sps1, sps2, threshold = 0.8)
print(result)


write.csv(result, file = "C:/Users/amentag/Desktop/these/These/clones/RDynlib/data/match_results.csv", 
          row.names = FALSE)
          
'''

```

