---
title: "Untitled"
format: html
server: shiny
---

## Shiny Documents

In this quarto document we calculate the similarity score of 4 spectra examples from the dynlib ftmsneg and flax seed ftmsneg databases using :

1.  dynlib_symmetric_dotproduct() function

2.  compareSpectra() from the spectra package

3.  NeutralLossesCosine() from the Python's matchms library

```{r}
library(Spectra)
```

```{r}
source("C:/Users/amentag/Desktop/these/These/clones/RDynlib/RDynLib/Similarity/sim_functions_with_options.R")
```

**test of real spectra using "dynlib_symmetric_dotproduct" function**

```{r}
  #first exemple (score = 0,99)
  x <- cbind(mz = c(129.456558227539),
             intensity = c(779.431457519531))
  y <- cbind(mz = c(129, 133, 159),
             intensity = c(339.653961, 119.266136, 133.141098))
  m <- dynlib_map(x, y, 177.041, 177.04)
  res <- dynlib_symmetric_dotproduct(m$x, m$y)
```

```{r}
 #second example(score = 1)
  x <- cbind(mz = c(128.455490112305),
             intensity = c(11648.4765625))
  y <- cbind(mz = c(128),
             intensity = c(741.387939))
  m <- dynlib_map(x, y, 257.078, 257.08)
  res <- dynlib_symmetric_dotproduct(m$x, m$y)
  
```

```{r}
 #third example(score = 1)
  x <- cbind(mz = c(159.619970320805, 315.01799205042),
             intensity = c(94.3498382568359 , 4343.77685546875))
  y <- cbind(mz = c(315, 433),
             intensity = c(3391.216064, 104.316986))
  m <- dynlib_map(x, y, 477.1023, 477.1)
  res <- dynlib_symmetric_dotproduct(m$x, m$y)
```

```{r}
  #fourth example(score = 0,99)
  x <- cbind(mz = c(300.446106481507),
             intensity = c(668.621826171875))
  y <- cbind(mz = c(179, 254, 255, 271, 272, 300, 301, 313, 316, 317, 343),
             intensity = c(150.978485, 105.706268, 507.930054, 433.778381, 293.672119, 5063.506348, 2164.525146, 124.878708, 308.779663, 837.373535, 281.371155))
  m <- dynlib_map(x, y, 463.03, 463.03)
  res <- dynlib_symmetric_dotproduct(m$x, m$y)
```

**test of real spectra using "compareSpectra" function**

```{r}
  #first example
  x <- DataFrame(precursorMz = 177.041)
  mz <- list(129.456558227539)
  intensity <- list(779.431457519531)
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)
  
  y <- DataFrame(precursorMz = 177.04)
  y$mz <- list(c(129, 133, 159))
  y$intensity <- list(c(339.653961, 119.266136, 133.141098))
  y <- Spectra(y)
  # the similarity score is 0.9559187 and 1 matched peak
 
  compareSpectra(
  x,
  y,
  ppm = 0,
  MAPFUN = function(x, y, ...) joinPeaks(x, y, tolerance = 0.5),
  FUN = function(x, y, ...) MsCoreUtils::ndotproduct(x, y, m = 0.6, n = 2),
  matchedPeaksCount = TRUE
)


  
```

```{r}
  
  #second example
  x <- DataFrame(precursorMz = 257.078)
  mz <- list(c(128.455490112305))
  intensity <- list(c(11648.4765625))
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)
  
  y <- DataFrame(precursorMz = 257.08)
  y$mz <- list(c(128))
  y$intensity <- list(c(741.387939))
  y <- Spectra(y)
  # the similarity score is 1 and 1 matched peak
  compareSpectra(
  x,
  y,
  ppm = 0,
  MAPFUN = function(x, y, ...) joinPeaks(x, y, tolerance = 0.5),
  FUN = function(x, y, ...) MsCoreUtils::ndotproduct(x, y, m = 0.6, n = 2),
  matchedPeaksCount = TRUE
)

```

```{r}
 #third example
  x <- DataFrame(precursorMz = 477.1023)
  mz <- list(c(159.619970320805, 315.01799205042))
  intensity <- list(c(94.3498382568359 , 4343.77685546875))
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)
  
  y <- DataFrame(precursorMz = 477.1)
  y$mz <- list(c(315, 433))
  y$intensity <- list(c(3391.216064, 104.316986))
  y <- Spectra(y)
  # the similarity score is 0.9999986 and 1 matched peak
  compareSpectra(
  x,
  y,
  ppm = 0,
  MAPFUN = function(x, y, ...) joinPeaks(x, y, tolerance = 0.5),
  FUN = function(x, y, ...) MsCoreUtils::ndotproduct(x, y, m = 0.6, n = 2),
  matchedPeaksCount = TRUE
)

```

```{r}
  #fourth example
  x <- DataFrame(precursorMz = 463.03)
  mz <- list(c(300.446106481507))
  intensity <- list(c(668.621826171875))
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)
  
  y <- DataFrame(precursorMz = 463.03)
  y$mz <- list(c(179, 254, 255, 271, 272, 300, 301, 313, 316, 317, 343))
  y$intensity <- list(c(150.978485, 105.706268, 507.930054, 433.778381, 293.672119, 5063.506348, 2164.525146, 124.878708, 308.779663, 837.373535, 281.371155))
  y <- Spectra(y)
  # the similarity score is 0.9666572, and 1 matched peak
  compareSpectra(
  x,
  y,
  ppm = 0,
  MAPFUN = function(x, y, ...) joinPeaks(x, y, tolerance = 0.5),
  FUN = function(x, y, ...) MsCoreUtils::ndotproduct(x, y, m = 0.6, n = 2),
  matchedPeaksCount = TRUE
)

```

**test of real spectra using "NeutralLossesCosine" function**

```{r}
#' R session:

install.packages("reticulate")
reticulate::install_miniconda()
```

```{r}
#' R session:

install.packages("BiocManager")


install.packages("remotes")
```

```{r}
#' R session:

BiocManager::install("RforMassSpectrometry/SpectriPy")
```

```{r}
#' R session:

library(Spectra)
library(SpectriPy)
```

```{r}
  #first example
  x <- DataFrame(precursorMz = 177.041)
  mz <- list(129.456558227539)
  intensity <- list(779.431457519531)
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)
  
  y <- DataFrame(precursorMz = 177.04)
  y$mz <- list(c(129, 133, 159))
  y$intensity <- list(c(339.653961, 119.266136, 133.141098))
  y <- Spectra(y)
  
  param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = TRUE
   )
  #the similarity score is 0.977711
  res <- compareSpectriPy(x, y, param = param)
  res
  
  
```

```{r}
  #second example
  x <- DataFrame(precursorMz = 257.078)
  mz <- list(c(128.455490112305))
  intensity <- list(c(11648.4765625))
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)
  
  y <- DataFrame(precursorMz = 257.08)
  y$mz <- list(c(128))
  y$intensity <- list(c(741.387939))
  y <- Spectra(y)
  
  param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = TRUE
   )
  #the similarity score is 1
  res <- compareSpectriPy(x, y, param = param)
  res
```

```{r}
 #third example
  x <- DataFrame(precursorMz = 477.1023)
  mz <- list(c(159.619970320805, 315.01799205042))
  intensity <- list(c(94.3498382568359 , 4343.77685546875))
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)
  
  y <- DataFrame(precursorMz = 477.1)
  y$mz <- list(c(315, 433))
  y$intensity <- list(c(3391.216064, 104.316986))
  y <- Spectra(y)
  
  param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = TRUE
   )
  #the similarity score is 0.9999993
  res <- compareSpectriPy(x, y, param = param)
  res
```

```{r}
 #fourth example
  x <- DataFrame(precursorMz = 463.03)
  mz <- list(c(300.446106481507))
  intensity <- list(c(668.621826171875))
  x$mz <- mz
  x$intensity <- intensity
  x<- Spectra(x)
  
  y <- DataFrame(precursorMz = 463.03)
  y$mz <- list(c(179, 254, 255, 271, 272, 300, 301, 313, 316, 317, 343))
  y$intensity <- list(c(150.978485, 105.706268, 507.930054, 433.778381, 293.672119, 5063.506348, 2164.525146, 124.878708, 308.779663, 837.373535, 281.371155))
  y <- Spectra(y)
  
  param <- NeutralLossesCosine(
  tolerance = 0.5,
  mz_power = 0.6,
  intensity_power = 2,
  ignore_peaks_above_precursor = TRUE
   )
  #the similarity score is 0.9831873
  res <- compareSpectriPy(x, y, param = param)
  res
```
